### 2021-01-27 manuscript.Rmd
Solid draft

---
title: Honor Among Thieves^[]
subtitle: How 19^th^ Century American Pirate Publishers Simulated Copyright Protection
date: \today
author: 
- Ryan Safner^[Hood College, Department of Economics and Business Administration; safner@hood.edu]

abstract: |
  | From 1790 to 1891, the United States prevented foreign authors from obtaining domestic copyright protection, implicitly subsidizing a domestic reprinting industry. With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce psuedo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection. This paper analyzes this system using the Bloomington School's Institutional Analysis and Design (IAD) framework to understand its effectiveness, and pitfalls, in managing the commons of unprotected foreign works in 19^th^ Century America.
  |
  | *JEL Classification:* H41, K29, N40, O34
  | *Keywords:* Copyright, intellectual property, economic history

bibliography: honoramongthieves.bib
geometry: margin = 1in
fontsize: 10pt
mainfont: Fira Sans Condensed
monofont: Fira Code
mathfont: Fira Sans
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_caption: yes
header-includes:
  # - \usepackage[utf8]{inputenc}
  - \usepackage{amsmath, setspace, multirow, natbib,url,booktabs,caption,setspace,rotating, tikz}
  - \usetikzlibrary{calc, shapes, arrows}
canonical: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\clearpage

# Introduction {#intro}

\doublespacing

> "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience...At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Spoo2013 [42]

From 1790 to 1891, the United States explicitly prevented foreign authors from obtaining domestic copyright protection for their works.[^1] This policy, in part, was a boon to the domestic publishing industry, which largely reprinted books by foreign authors for an American audience [@Safnerpiratethyneighbor]. American publishers at the time, deemed "pirates" by their detractors, treated unprotected foreign works as a commons. As the United States developed and publishers began to "print on" each other (by reprinting each other's reprints of foreign works), the publishers were forced to develop a system to prevent the foreign-work commons from being depleted from ruinous competition and price wars. This system entirely composed of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid to the late 19^th^ century when the U.S. began the process of granting copyright protection to foreign authors.

[^1]: Even after the 1891 Chace Act that first gave copyright protection in the U.S. to foreign authors, it was *de facto* quite costly for authors to attain this. Additional requirements, some overtly protectionist such as the "manufacturing clauses" requiring foreign works to be printed in the U.S. by U.S. printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention standardizing international copyright until 1989.

This paper examines the effectiveness of the courtesy of the trade system as a strategy by publishers to manage the common pool resource of unprotected foreign works. I use the Bloomington School's IAD framework to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulate the procedures and protections that actual copyright protection brought. While it was not uniformly followed by all publishers, saw occasional breakdown into price wars, and ultimately faded with the formal recognition of foreign copyrights, it serves as an interesting example of both how informal norms can resolve disputes, as well as an alternative mechanism of protecting intellectual property without statutory law.

Economists typically define a commons, or a common pool resource, as being rivalrous but non-excludable: each person's consumption subtracts from another's ability to consume, but all such users are unable to prevent others from accessing the resource. @Hardin1968 famously termed the situation a "tragedy of the commons," as the inability of users to exclude others' use logically implies users will rapidly deplete the resource. A classic solution, among economists, has been to establish property rights over the resource, allowing owners to exclude other users, internalizing the benefits and responsibility to improve and stewarding over the resource [@Alchian1965; @Demsetz1964]. However, establishing and enforcing property rights is costly, and are only done when the net benefits of "propertization" exceed the net costs of maintenance (or the net benefits of a commons) [@Demsetz1969]. Indeed, economists have shown that if rights are designed poorly and allocated from above, then a symmetric tragedy of the *anti*-commons may result, where any one possessor of exclusion-rights among many can veto the productive use of a resource [@Heller1997; @Buchanan2000; @Heller2008].

Research by Bloomington School have focused on the importance of informal institutions in managing a commons, cataloging a series of cases and developing a rich grammar of institutions in the Institutional Analysis and Development (IAD) framework [@Ostrom1990; @Hess2007; @McGinnis1999; @Cole2014].

Many researchers focusing broadly on the topic of culture, expression, and knowledge, have productively analyzed these concepts as a commons [@Benkler2002; @Hess2007; @DouradoTabarrok2015; @Schweik2012; @Frischmann2014]. @Hess2007 [3] take a wider interpretation of commons, defining one as "a resource shared by a group of people that is subject to a social dilemma." [@Ostrom1990,90; @Hess2007,7; @Wilson2013,S22] summarize eight common principles among institutions that successfully manage a commons, paraphrased below:

1.  Clearly defined boundaries for group membership and for the shared resource
2.  Proportional equivalence between benefits/costs and the contributions/transgressions of members
3.  Collective-choice arrangements to allow members to establish rules and make decisions for the group
4.  Monitoring of member behavior through detectable norm-abidement to prevent free-riding
5.  Graduated sanctions for transgressors ranging from informal gossip to expulsion
6.  Conflict-resolution mechanisms that are viewed as efficient and fair
7.  Recognition of rights of group members to self-organize internally
8.  Polycentric relations between the group and other social orders to maintain optimal size and autonomy

@DouradoTabarrok2015 call this the "eightfold path to success" and apply this to ideas and inventions, which they see as a "Super-Lockean commons" which, if accessible, actually *grows* the more that people productively use it (c.f. @Hess2007,5). @Safner2016 applies these 8 principles to explain the success of Wikipedia, the free online encyclopedia, in managing public knowledge.

Federally-sponsored American literary piracy has been noted and extensively discussed by lawyers, historians, and literary scholars. Many have examined the legal development of American (and previously, British) copyright laws, noting the lapse in protection to foreign authors in the United States throughout the 19^th^ century (among the most systematic are @Patterson1968 and @Johns2009). A number have also examined the movement for international copyright in the United States that achieved partial victory at the end of the 1880s (see e.g. @Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).

While there is a vast literature in the economics of copyright and intellectual property more broadly,[^2] economists have largely overlooked the dynamics of the event except in the service of broader arguments or economic history regarding intellectual property rights. @Khan2005, however, recognizes the idiosyncratic and endogenous component of copyright history, primarily to distinguish the unique and superior American system of IP from that of European systems. @Khan2004 examines its welfare effects on American authors, publishers, and consumers, finding it a net positive. @Spoo2013 [Ch.1] notably places the workings of the courtesy system in the literature of solutions to the tragedy of the commons, citing touchstone works such as @Ellickson1994 on neighbors solving disputes and @Rustiala2012 on norms in fashion, comedy, and cooking. However, the system has yet to be explicitly analyzed in an IAD framework to explain its origin, mechanisms accounting for its success, and decline, as an institution. This paper examines how the set of norms and practices constituting "trade courtesy" in the 19<sup>th</sup> century adhere to the 8 principles of success outlined above, as well as how they led to the emergence and eventual decline of this system of self-regulation.

[^2]: See e.g. @Besen1991; @Posner2003 for a good survey. For a growing corpus of more critical works, see e.g. @Boldrin2008; @Kinsella2008; @Heller2008; @Rustiala2012; @Bell2014; @DouradoTabarrok2015

# 19^th^ Century American Literary Piracy and Courtesy of the Trade

> "[A] literary pirate is not only not an outlaw; he is protected by the law. He is the product of law." [@PW1882, 430]

> "[Trade courtesy] was a brief realization of the ideals of philosophical anarchism--- self-regulation without law. ---Henry Holt (1908)2

Statutory copyright came to the United States with the Copyright Act of 1790, its authority derived from the Copyrights Clause of the U.S. Constitution[^3] [@Pattersonbook]. Largely mimicking the 1710 Statute of Anne in Britain - the first modern copyright statute - it granted "sole right and liberty of printing, reprinting, publishing and vending" to authors of "any map, chart, book or books already printed within these United States" for "the term of fourteen years from the recording the title thereof in the clerk's office," with an optional 14-year renewal term (Copyright Act of 1790, \S 1). Crucially for American publishers, section 5 of the Act explicitly denies copyright protection to foreigners:

[^3]: "To promote the Progress of Science and useful Arts, by securing for limited Times to Authors and Inventors the exclusive Right to their respective Writings and Discoveries" (U.S. Constitution, Article I, \S8, Cl. 8).

> "And be it further enacted, That nothing in this act shall be construed to extend to prohibit the importation or vending, Reprinting or publishing within the United States, of any map, chart, book or books, written, printed, or published by any person not a citizen of the United States, in foreign parts or places without the jurisdiction of the United States" (U.S. Copyright Act of 1790, \S5).

The domestic publishing industry in the early United States was largely run by Scottish and Irish immigrants such as Mathew Carey, who transplanted their tried-and-true experiences pirating London's books in Edinburgh and Dublin into New York and Philadelphia [@Johns2009, p.188ff; @Spoo2013, 34; @Groves2007b, 141].[^4] Early politicians and pundits saw reprinting as key to the emerging "American system" of political economy championed by Alexander Hamilton, Henry Clay, and Friedrich List, whereby tariffs and other measures would protect infant manufacturing industries. Thinkers of this school saw the reprinting (authorized or not) of European (predominantly British) works as key to instilling mass literacy, inspiration for innovation, instilling republican virtue [@Johns2009; @Spoo2013, 28; @Clark1960, 70]. In the first few decades of the American colonies and early republic, the great distances between major cities kept parochial printing presses from cannibalizing each's hold over their local market.

[^4]: Prior to the 1707 and 1801 Acts of Union --- expanding the English crown and Parliament's sovereignty over Scotland and Ireland, respectively --- publishers in Edinburgh and Dublin had thrived off the lack of copyright protection on English works in their respective countries. To avoid Scottish/Irish publishers printing on other Scottish/Irish publishers, they set up courtesies like those described in the American context, to maintain a commons over unauthorized English reprints. Upon integration into the United Kingdom, copyright protection established in the 1710 Statute of Anne compelled these publishers to turn "legitimate" or take their business model to the American colonies.

At mid-century, there were about 400 publishers, with a supermajority of them in New York, Philadelphia, Baltimore, and Boston, according to the authoritative *Trubner's Bibliographical Guide to American Literature*, compiled in 1859, [@Shove1936, 23]. Harper & Brothers was the largest firm, with over 2 million volumes sold in 1859.

The growth of the industry brought about fierce competition, especially after the Depression in the late 1830s and a ruinous price war into the 1840s. The publisher Carey & Lea, for example, had seen their output of new titles shrink by over 50% between 1835 and 1837 [@Groves2007b, 141-142]. Publishers also resorted to clever alternative forms of competition, such as the "weeklies" which printed entire English novels on "mammoth" size single sheets sent through the mail direct to customers, akin to newspapers [@Groves2007b, 142; @Barnes1974, 4-24]. These changes underscored the need for self-regulation and organizational solutions within the industry into the 1840s. Some publishers, such as Carey, suggested forming an explicit organization modeled off of the London Stationers Company. Carey set up The Philadelphia Company as an explicit cartel arrangement, only to see it evaporate 5 years later.[^5] Despite its failure, @Johns2009 [199] summarizes the common sentiment within the publishing industry:

[^5]: Once such a cartel allocated titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint. Without an inflexible cartel, a bargain could be reached between two publishers to establish boundaries, but this was not possible with a large organization and rogue outside printers.

> "What all such suggestions shared were three convictions: the paramount importance of reprinting European works; the consequent need to eliminate domestic reprinting and rival importing (in their terms, piracy); and the requirement that a solution to these problems come from the trade itself."

Instead, publishers slowly developed a set of norms, commonly referred to as "trade courtesy" or "courtesy of the trade" to prevent domestic publishers from "printing on" one another, as they themselves printed on foreign works. While foreign authors and printers surely denounced most American publishers as "pirates" for their (frequently unauthorized) reprinting of foreign works in America, the American publishers reserved the epithet of "pirate" for those presses who refused to abide by trade courtesy. Publisher and famed champion of trade courtesy, Henry Holt, contrasted the "outside barbarian" with the "men of exceptional character" who observed courtesy @Spoo2013, 50]. Indeed, as the system of courtesy was nearing its demise, Samuel Clemens (pen name Mark Twain) testified before the U.S. Senate in 1886 about the nature of "piracy":

> "I do consider that those persons who are called 'pirates'...were made pirate by the collusion of the United States Government...Congress, if anybody, is to blame for their action. It is not dishonesty. They have that right, and they have been working under that right a long time, publishing what is called 'pirated books.'" quoted in @Spoo2013 [22].

The remainder of this section explores how the system of trade courtesy worked, by examining how it exemplifies the 8 design principles identified by Bloomington School research. In the process, I explain how the set of practices originated and matured, and what caused its demise by the 20<sup>th</sup> century. Table 1 summarizes some key events that shaped the history of the courtesy system to provide additional context.

```{r, echo =F, message=F, warning=F}
library(tidyverse)
tribble(
  ~"Date/s", ~"Event",
  "1790", "Copyright Act exempts foreign authors from copyright protection in U.S.",
  "1800-1801", "Acts of Union establishes copyright protection of British authors in Ireland",
  "1816", "First tariff on imported books",
  "c.1830s-1840s", "First price war",
  "1850s", "More explicit formulation of courtesy",
  "c.1870s-1880s", "Second price war",
  "1891", "Copyright Act grants foreign authors copyright protection in U.S."
) %>%
  knitr::kable(format = "latex",
               caption = "Outline of Courtesy Period in United States",
               booktabs = TRUE)
```

## Defined Boundaries

The foreign reprinting industry in the United States was spawned directly from Section 5 of the Copyright Act of 1790. An American publisher could publish works by *American* authors only by acquiring the copyright from them first. Since section 5 exempted *foreign* authors from copyright protection, it provided an indirect subsidy to reprint these works without concern for acquiring valid copyrights. As such, rather than paying and haggling with an *American* author for her copyright, an American publisher could print European authors at a significantly reduced cost. Both established and upstart publishers rapidly entered this market, searching for a quick opportunity to republish a "sure thing" to an American populace hungry for English literature.

Such a nearly "free" resource, however, could easily see its value depleted as American publishers began to "print on" other publishers and lowering the price of a particular work in great supply.[^6]

[^6]: Ironically, as described below, as part of the system that emerged to prevent this very outcome, many publishers actually would end up *paying* considerable amounts to foreign authors to maintain priority in republishing their works in America.

The informal "courtesy of the trade" system of norms was both voluntary and ever-changing with market conditions. Not all publishers abided by trade courtesy, and some that followed norms closely for one decade would later violate them when it was to their benefit later. Some firms that had cut their teeth in the profession by reprinting other publishers' reprints of foreign works later turned "legitimate" by following trade courtesy and enforcing it against upstart pirates.

The objective of trade courtesy was to create a process by which pseudo-property rights could be acquired, transferred, and violations be remedied in a way that would be recognized by enough participants to be reasonably self-enforcing. As such, the *raison d'etre* of the system was to define the boundaries over to what did and did not constitute a publisher's "property," and to resolve disputes over those boundaries.

By extension of the existing principles, publishers following trade courtesy also developed an "options" system for firms that had previously published a specific author's work, whereby other firms would generally recognize that publisher as being "associated" with the author, and therefore obtain first claim on new works by that author. This was referred to in the trade by the "rule of association" [@Spoo2013,38]. Aside from providing a bright line rule for priority of claims, a major benefit of the association rule was the a firm could advertise themselves as being the "authorized" publisher of a foreign author [@Spoo2013,42].

However, simply asserting that one *claimed* first priority in publishing an author was insufficient, and could incentivize "courtesy creep," [@Spoo2013,37]: Publishers might announce simultaneously or simply announce their intent to print any title that sounded promising, hoping to establish priority even if they would never print. Such a "first to announce" rule could quickly lead to inefficient rent-seeking: claimants merely assert claim in order to block others from using it, even if the claimant never intends to develop the resources in a productive way [@Posner2003,17]. Such a rule risks creating an "anti-commons" where multiple firms would seek to block any one firm from production [@Heller1997; @Heller1998; @Heller2008].

Instead of a "first to announce rule," a "committed-searcher rule," where the first claimant to undertake costly steps (strongly correlated with actually developing the resource) improves upon these incentives [@Posner2003,17]. Publishers following trade courtesy required a firm seeking to stake a claim to undertake a more costly action, often reaching a public agreement with the foreign author, or more frequently, purchasing advance sheets of the edition abroad. Holt summarizes: "If a publisher had the advance sheets in his possession, such right or claim overrode a simple announcement," (quoted in @Spoo2013, 37). @Spoo2013 [37] further describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown, and the firm undertaking a risky experiment.

## Proportional Costs & Benefits

In order for sustainable management of a commons, users must obtain benefits proportionate to the costs they bear. Otherwise, if a substantial subset of the group free rides without contributing to the commons, or contributes without any reward, the tragedy of the commons may rear its ugly head, and the resource will be depleted. In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs.\
\
The primary mechanism for upholding the system of norms and preventing tragedy was reciprocity between publishers --- publishers would respect the pseudo-property rights of other publishers to particular works in the hope that others would respect their own claims. To reiterate Holt's message, "anybody is welcome who will behave himself."

One of the main currencies in the publishing trade was reputation. Indeed, the more established publishing houses, which often upheld trade courtesy, saw it as a "gentlemanly" practice, almost the duty of any respectable publisher. Holt claimed that none of the major publishers[^7] "would go for another's author any more than for his watch." [quoted in @Spoo2013,40].

[^7]: @Spoo2013 lists Putnam, Appleton, Harper, and Scribner fitting this description.

Naturally, upstart publishers that were only in it for a quick buck had little incentive to worry about reputation. So long as there were not too many publishers competing over price in this way, the industry could maintain this equilibrium.

Upholding proper "gentlemanly" conduct was not costless, it did require publishers to sacrifice profitable opportunities in the hope of not destroying their reputation (and hopes of future profit opportunities). As mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher $A$ seeking to reprint works of foreign author \$X\$ (and hoping to beat publisher \$B\$ to the punch) could not merely assert this and hope publisher \$B\$ would back down; publisher \$A\$ needs to strike an exclusive arrangement with author \$X\$ or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.

This required investment led to one of the great ironies of the system: despite the absence of a legal duty, U.S. publishers frequently paid significant honoraria to British authors for the "right" to print their works. In fact, publishers often *emphasized* in their advertisements the amount they paid to foreigners for their "exclusive" or "authorized" edition [@Spoo2013, 24]. Occasionally, it would be payments to the British publisher for advance sheets. Harper & Brothers, for example, paid Edward Bulwer-Lytton £50 per volume, Charles Dickens between £250 and £2,000 for his novels [@Spoo2013, 39]. Carey paid Sir Walter Scott £75 for advance sheets of each of the Waverley novels and £300 for *Life of Napoleon Bonaparte*. The Harpers described the courtesy practice of purchasing advance sheets from foreign authors or their agents or publishers:

> "In the absence of an international copyright, it is the custom for an English author, or his agent in London, to send early sheets to some American publisher, fixing a price therefor, and by a law of courtesy the American publisher who has issued the previous works of an author is entitled to the first consideration of that author's new book" (quoted in @Spoo2013,39).

Henry Holt, in his 1893 reflections on trade courtesy, notes that "It not only prevented ruinous competition between American publishers, but also secured to foreign authors most of their rights," (ibid, 32). This practice gave an aura of legitimacy to the courtesy system, at least before members of the public (but certainly not authors). The *American Publishers' Circular and Literary Gazette* boldly declared in 1872 that it was "an undeniable fact that there is no living English author of established reputation, whose books are extensively republished in this country, who is not freely and properly compensated by the American publisher," (quoted in @Sheehan1951,68). George Putnam, more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," (quoted in @Spoo2013,41). By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146]. \
\
Of course, the price of exclusivity on the most popular novels and authors was prohibitively high. The temptation to pirates of printing a "sure thing" was too great for any courtesy-abiding publisher to purchase exclusivity with any expectation of it being respected [@Groves2007b, 144]. Similarly, "classic" (often dead) authors and collected works were unable to be associated with any single publisher.\
\
In general, the cost of a publisher creating an association with a foreign author under trade courtesy was proportionate to the benefits of expecting others to refrain from printing on one's works. These pseudo-property rights and attendant norms regarding their establishment and protection ensured that rule-abiding publishers could expect stability in their business and stave off a price war, so long as they bore some of the cost of upholding the system.

## Collective-choice Arrangements

Any successful commons must have some set of decision-making procedures to create and modify rules affecting the users in order to further the group's ultimate objectives. For the American reprint publishers, this was always to avoid conflict and ensure an orderly system of pseudo-property rights. Much like (legitimate) Anglo-American property law and the merchant-made law of *lex mercatoria* [@Benson1989], the set of rules was an emergent result of judgments in adversarial disputes. It was a system neither designed or managed by any organization, nor could any single publisher directly steer it in any particular direction. Chronicling the history of the corpus of practices, Holt comments: "Trade courtesy is as full of exceptions as the law itself. It has grown up as a mass of decisions in particular cases, just as the common law has," (quoted in @Spoo2013, 38).

The main "objective" of the system, only insofar as all members shared such a goal, was merely to not dilute the commons. The fear of a ruinous price war loomed in the back of the mind of all publishers engaged in active disputes with one another if negotiations broke down.

Over the century, some publishers occasionally attempted to organize into more formal and explicit associations or cooperate as a large corporation, in order to rationalize the industry. Some, like Henry Carey, not-so-secretly dreamed of organizing an American guild like the Stationers Company back in England. However, such a scheme always failed because it lacked the flexibility of the courtesy system in two important ways [@Johns2009,197-202]: First, such a corporation would generate a list of titles (akin to the Register of the Stationers Company) available only to its members --- boosting a signal to rogue reprinters that these titles were "sure things" worth printing on. Second, a formal corporation prevented the very bargains and exchanges between publishers to settle psuedo-property rights disputes to prevent American publishers printing on each other.

## Monitoring

Any set of norms or institutions for managing a commons requires diligent monitoring of users by other users, to ensure that no user is free riding and failing in their duties of stewardship.\
\
While trade courtesy norms could not sustain themselves without vigorous monitoring, each individual publisher had a strong incentive to monitor others for violating norms. If a publisher's works were printed on by other publishers, that publisher stood to lose profits from competition. Thus, each individual publisher had an interest in ensuring that other publishers respected their claims, and via reciprocity in this manner, the system of trade courtesy was largely self-enforcing.\
\
However, it is still instructive to see how publishers commonly took several types of actions in order to notify others of their claims and dispute the claims of others in a public manner.\
\
Since the early days of trade courtesy, the publishers were often "men of letters" and frequently wrote to one another and occasionally published their correspondence for others in their network to see. This network proved critical for communicating about claims, gossiping about others' claims, resolving honest disputes, and calling out transgressors. Publishers regularly wrote to their customers, advertised in newspapers, and created a number of trade journals to circulate announcements, reviews, and notices to one another [@Groves2007a]. Periodicals such as *Publishers' Weekly*, which by the 1880s emerged dominant within the trade, also advocated consensus positions within the trade for regulatory reform.

Beyond the frequent communication between individual publishers, it was essential to maintain the nexus of assertions of ownership and association with foreign authors publicly, by posting various messages in widely-read trade journals. In the 1850s, there was no agreed upon single medium, as various publishers utilized different trade journals and weeklies. Due to the lack of a single medium, publishers frequently clipped and kept records of their own advertisements in media in order to provide justification to their claims should disputes later arise @Groves2007b, 142]. As the practice matured, publishers eventually reached consensus that claims "in press" should be advertised and declared in the *New York Commercial Advertiser* [@Groves2007b, 146; @Spoo2013, 37; @Sheehan1951, 69]. Other major trade journals such as *The Publishers' Weekly* and *The Weekly Trade Circular* then reprinted the advertisements from the *New York Commercial Advertiser* to ensure sufficient dissemination that publishers in other cities were properly warned not to print on the claim. An example letter from Scribner, Armstrong, and Co. to Roberts Brothers expresses:

> "We notice in last evening's Commercial Advertiser your 'in press' announcement of the 'Life of Marie Antoinette' by C. D. Yonge. We beg to call your attention to our previous announcement (March 29th) and to an early copy of the work in hand. We hope therefore that you will not interfere with us in the matter," quoted in @Sheehan1951,69.

## Graduated Sanctions for Transgressors

In order for any commons to avoid tragedy, free riders and transgressors must be deterred and punished for abusing the system, proportionate to their transgressions. Furthermore, punishments should be graduated, to ensure that there is an ever-increasing marginal cost of successively violating more important norms.

As there were no legally-recognized property rights where an injured publisher could take a violator to court, disputes needed to be settled between aggrieved parties, or by "appeal" to other publishers, all of whom have an interest in maintaining courtesy norms.

An earnest dispute over who had priority might first be met with appeals to publishers' reputation or public shaming. @Spoo2013 [43] cites an example in 1870 where the Harpers attempted to print on Holt's edition of Hippolyte Taine's *On Intelligence*. Holt wrote to the Harpers merely asking "Doesn't the fact that we have published several of [Taine's] books entitle us to [exclusively publishing \*On Intelligence\*] if we want it?" Such a simple response was sufficient for the Harpers to back off, abiding by the rule of association. Upon similarly resolving a second dispute between the firms, Holt graciously expressed that Harper had followed "what the notions of honor...prevalent among publishers of standing required," (ibid).

Additionally, publishers might turn to the reading public, in hopes that market competition would sort out who was the more "authentic" publisher. Firms that were associated with foreign authors might protect their claim against other publishers printing on them by including copies of letters penned by the foreign author at the beginning of the book. Ticknor & Fields editions of Thomas De Quincey's works contained a letter featuring the latter approving of the former "exclusively" publishing his works in America, as did United States Book Co. demonstrating they obtained by Rudyard Kipling's "authority," the same for his *Mine Own People* [@Spoo2013,42].

Publishers in dispute might also turn to the trade press, appealing to their peers, with evidence that they had priority due to the rules of association. Harper Brothers placed a full-page advertisement in *Publisher's Weekly* documenting all of Thomas Carlyle's works they had published, in order to claim priority in publishing his *Reminiscences* [@Spoo2013,44; c.f. @Sheehan1951,71]. A rival publisher, Charles Scribner's Sons, responded with their own full page advertisement documenting exclusive arrangements they had with Carlyle's niece in obtaining advanced sheets, asserting, "The public will choose between this edition, put forth by the clearly expressed authority of Mr. Carlyle's executor, and a reprint from our sheets [by Harper] under a claim to which he has distinctly refused his acknowledgment," [@Spoo2013,44]. Ironically, one of the most effective forms of shaming was to denounce transgressing publishers as "pirates" before their more respectable peers. At a libel lawsuit reported by *Publisher's Weekly*, Holt and other major publishers testified that use of the term "pirate" had a special meaning in American publishing - as a public denunciation within the profession for those that refused to abide by trade courtesy [@Publisherslibel]. If a publisher were recalcitrant to such public denunciations, publishers could rally together for other multilateral punishments, such as refusal to deal with a rogue publisher.\
\
The ultimate sanction against a "pirate", which if not kept in check by agreement that its *occasional* use was legitimate, was retaliation by predatory pricing. While the *raison d'etre* of trade courtesy was to keep prices sufficiently high to avoid a price war, publishers recognized it as legitimate as a response or a preemptive strike against pirates. If firm $A$ prints on firm \$B\$ and the latter has a clear claim to priority, firm \$B\$ might reissue the same work below cost to drive firm \$A\$ out of the market (at least for that title). Joseph Henry Harper, of Harper & Brothers wrote:

> "If a publisher declined to comply with the requirements of trade courtesy, some method would be adopted to discipline the offender---generally by the printing of lower-priced editions of his foreign reprints by his aggrieved competitor" (quoted in @Spoo2013,46).

\
Print on other publishers' claims ran frequent during the two price wars of the 19<sup>th</sup> century. In the 1870s, for example, Harper Brothers reprinted their previously published foreign novels at just 10 cents a piece. The Harpers made their intentions clear in an 1879 letter:

> "We determined that [the cheap reprinters] should not share our profits, because we intended that there should be no profit for a division. We began to print on ourselves" [quoted in @Spoo2013,46].

\
Upstart firms were less able to withstand below-cost prices than established publishing houses. While the established publishing houses would lose profits, and viewed printing on those who printed on them as a last resort, they could withstand it longer than upstart firms.

However, even the most respectable and established publishers saw this as a last resort. In response to the breakdown in negotiations between Harper and Scribner over the rights to the late Thomas Carlyle's *Reminisciences* mentioned above*,* Harper printed on Scribner. Scribner, whom many publishers saw as having the legitimate "right" to retaliate by printing on Harper, refused to "descen[d] to blatant piracy" and sought to maintain the moral high ground in the eyes of British writers [@Spoo2013,47; @Sheehan1951,71]. Similarly, Holt, in a letter to George Putnam in 1880, declared:

> "I don't believe, under any ordinary circumstances, in the reprisal policy, and there's no form of stupidity that I so thoroughly detest (whether I am exempt from it or not) as that which attempts to make its own success out of successes legitimately belonging to other people. Nowhere that I know of does intelligent selfishness so much consist in altruism as in American publishing," quoted in @Sheehan1951 [73].

Retaliation, if not kept in check, could easily spin out of control. When Harper & Brothers failed to resolve a dispute over Tennyson's works with Fields, Osgood & Company --- who had been associated with the author for 30 years --- in 1870, Harper released their own edition [@Sheehan1951,145]. The blood in the water encouraged other publishers to release their own editions of Tennyson, all but evaporating Fields' longstanding claim. @Sheehan1951[145] comments, "once a publisher lost control of an association, it could not be reclaimed--- almost as if a copyright had expired and a book had slipped into the public domain."

## Conflict-resolution Mechanisms

Before a dispute need devolve into the series of sanctions discussed above, successful management of a commons requires mechanisms for impartial adjudication. Before resorting to the punishments described above, publishers competing over a claim often employed several different methods to attain an agreeable resolution: compensation and "adjustments," and third party arbitration.

As discussed above, publishers were in frequent contact with one another via letters, as well as messages broadcasted to all in trade journals. A simple first response to a publisher printing on another was for them to write to one another and reach an informal settlement, or an "adjustment" that often took the form of monetary or in-kind compensation.

@Sheehan1951 [68] recounts several cases of adjustments: Rudd & Carleton sought to publish an English translation of Alexander von Humboldt's correspondence in 1860 without realizing that Appleton had already purchased advance sheets in English, intending to print. Appleton gave up the volume to Rudd & Carleton in exchange for the £40 the latter had already invested in the project. Similarly, Ticknor & Fields and Harper found themselves in a dispute over issuing Charles Dickens' *Mystery of Edwin Drood*, ultimately agreeing that Ticknor & Fields would issue the novel as a single book and Harper would bring it out serially in *Harper's Weekly* [@Spoo2013,48]. A common solution was for one claimaint to issue an edition in (relatively more expensive) cloth, and another claimant to issue the same work in cheaper paperback; alternatively, if the work were one of a series, one claimant release the disputed volume, and another release the following volume, @Sheehan1951 [74]. Much like the owner of a true copyright today, a publisher with a good claim to issuing a foreign work in the U.S. could bargain with other publishers over a wide range of in-kind rights - issuing serials, hardbound books, paperback books, etc. - to reach an amicable arrangement.\
\
Publishers that were unable to reach bilateral agreement often submitted the competing claims before an agreed upon third publisher to act as a neutral arbitrator. Harper describes how "the contention being commonly left to a fellow-publisher for arbitrament," (quoted in @Spoo2013, 48). Correspondance of 40 letters sent by Charles Scribner's Sons to 9 firms[^8] in the 1870s suggest most of these disputes were amicably resolved [@Sheehan1951,70].

[^8]: J.B. Lippincott and Co., J.R. Osgood and Co., D. Appleton and Co., Roberts Brothers, G.P. Putnam's Sons, Harper and Brothers, Macmillan and Co., E.P. Dutton and Co., and Henry Holt and Co.

## Recognition of Self-organization Rights

Courtesy of the trade was always a voluntary code of norms, slowly internalized over the course of the 19^th^ Century. Throughout that period, as described above, various publishers had attempted (and largely failed) to organize trade associations or explicit cartel arrangements, in addition to two periods of fierce price competition between new entrants and established publishers.\
\
It was common during the price wars, especially the latter one in the 1870s and 1880s, for upstart firms to enter the market and refuse to abide by trade courtesy [@Spoo2013,47]. A new class of pirate publishers, like John W. Lovell, Isaac K. Funk, and George P. Munro, emerged with competitive innovations in the 1870s --- creating low quality, cheap "libraries" of reprinted works sent through the mail at subsidized second-class rates [@Groves2007b, 146]. In 1886 alone, 26 "cheap libraries" issued 1,500 different titles, mostly novels priced between 10-50 cents, thinning the industry's profit margins [@Spoo2013,53]. Appealing to populist notions of cheap literature for the masses, these entrants directly attacked courtesy as a price-fixing cartel. Lovell opined in an 1879 edition of *Publishers' Weekly* that:

> "I can say to the younger and smaller [publishing] houses from my own experience, go in heartily for the 'courtesy of the trade' and---starve. You will find everything is expected of you and very little given you. As for my part, I prefer to follow the examples that led to success in the past rather than the precepts now advocated to prevent others from attaining it," [quoted in @Groves2007b].

Munro declared that "cheap libraries have broken down the Chinese or rather American wall of trade courtesy and privilege [erected solely for the] monopoly of publishers in this country," [quoted in @Spoo2013, 53-54].

@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^9] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914. Furthermore, publishing firms, if they were incorporated at all, tended to be closely-held by family outfits.\
\
Furthermore, even those firms that upheld trade courtesy knew it was imperfect, and was little more than a paper shield against determined pirates. Publishers chief weapon against pirates was speed [@Groves2007b; @Johns2009; @Spoo2013; @Sheehan1951]. As early as 1823, a letter from Carey in Philadelphia to Sir Walter Scott's Edinburgh publisher boasts:

[^9]: However, this is not necessarily true of certain segments of the book market. @Sheehan1951 [50-51] describes the origins of the *American Book Company*, the result of a merger of several leading textbook-publishing firms in the late 1880s, which was estimated to control somewhere between 50-90% of educational textbooks, sufficient to garner antitrust attention by 1909. This market, however, is beyond the scope of this paper, which focuses on uncopyrighted foreign works.

> "We have [the] Game completely in our hands this time. In 28 hours after receiving [a copy of Scott's*Quentin Durward*], we had 1500 copies off or ready to go...In two days we shall publish it here and in New York, and the Pirates may print it as soon as they please. The [pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Spoo2013, 50].

Publishers hired foreign agents in London to obtain intelligence about what firms were to print which British books, and to set up relationships with authors and publishers to safely shuttle works and materials to their press in the U.S. before others were able to print. Even a handful of hours made all the difference between a profit or a pirate-induced loss on a venture.

## Polycentric external relations

Trade courtesy in publishing was primarily a response to explicit legal institutions, particularly the gap in copyright protection between domestic and foreign authors. Historical trade practices in printing, since the dawn of the printing press, had been shaped by changes in legal and political institutions. Since the first printers in England in 1476, printers had tried to organize themselves internally to manage the commons of "printing on" each other's works, regardless of whether the legal environment recognized explicit property rights in works. Many of the trade practices originate in the 16^th^-18^th^ century London printers, who had initially established a guild---the Stationers' Company---to organize their industry. Like most medieval craft guilds, the English crown granted the Stationers the sole right to print within the realm, and for them to govern their own trade. While often politically controversial in the tumultuous English Civil Wars of the 17^th^ Century [@Safnerpatronage], the main feature of the Stationers' practices was the establishment of a Register that maintained whom had the exclusive, perpetual, and transferable right to print a particular work (known as "a copy right") as well as a series of sanctions and methods of adjudicating disputes based on copies authorized by the Register. Alongside this grew up "trade customs" of printing as an honorable craft. The passage of the Statute of Anne in 1710 created the first statutory copyright in th U.K. (a temporary legal monopoly), but left unaddressed the status of the Stationers' copyright until the 1774 case of *Donaldson v. Beckett*, when the House of Lords decided that literary expression is protected only by statutory copyright.[^10] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy of famous works out of London. As noted above, many of these tradesmen emigrated to the early United States to carry on their esteemed tradition of pirating English works.

[^10]: The Stationers conjured a clever rhetorical idea that that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the protections of the Statute of Anne, a legal claim the Lords definitely rejected, @Johns2009.

The courtesy of the trade was only one aspect of a whole political economy of the young republic, a significant part of which revolved around protecting American publishers and readers. Many argued, especially in the early 19^th^ Century for a new "American system" of mercantilism favorable to domestic manufacturing in order to promote national development. The "appropriation" of European literature and technology was encouraged as a stimulant to American development and even instilling republican virtue [@Johns2009, 203]. A gradually-increasing tariff (ranging from 10-25%) slowed the importation of foreign books starting in 1816 and implicitly subsidized the American publishing (and reprint) industry [@Groves2007b, 141; @Dozer1949, 73]. @Dozer1949 [95] argues that the issue of respecting foreign copyright in America "has been...a tariff question involving the protection of American manufacturing interests." A federal court also recognizes the 1909 Copyright Act and a later Tariff Act (of 1930) as sharing a common purpose, "to encourage the industries of the United States," and "to protect American labor," (*Oxford University Press v. United States*, 1945 (No. 4491) at 20; quoted in @Spoo1998,70).

Of course, while the courtesy of the trade was able to often replicate the features of property rights and even copyright law itself, such voluntary norms were not viewed legitimate or enforceable property rights in court. The federal court in the 1865 case of *Sheldon v. Houghton*, 21 F. Cas. 1239, 1241--42 (C.C.S.D.N.Y. 1865)[^11] recognized this fact:

[^11]: Sheldon & Co. and Houghton & Co. had agreed in 1861 to jointly publish an edition of Charles Dickens' works. When Houghton withdrew in 1865, Sheldon sued for a breach of contract, citing a significant reliance was created in the partnership, according to the protections established by the trade courtesy system. Interestingly, both parties' attorneys obtained testimony of major publishers about the workings and legitimacy of trade courtesy. This is recounted in @Groves2007b[145].

> [trade courtesy] confessedly rests upon no common law of the country, recognized and administered by judicial tribunals. If it has any foundation at all, it stands on the mere will, or...the "courtesy" of the trade...It can, therefore, hardly be called property at all---certainly not in any sense known to the law, (quoted in @Spoo2013, 34).

The system rapidly broke down after 1891 with the passage of the International Copyright Act.[^12] The Act recognized foreign copyrights in the United States provided that the works themselves were manufactured in the United States. This notorious "manufacturing clause," as well as other qualifications and requirements kept the United States ineligible from joining the 1886 Berne Convention until 1989, following major revisions to U.S. law. U.S. Register of Copyrights Barbara Ringer comments that the law had enough loopholes to "make the extension of copyright protection to foreigners illusory" [@Ringer1968,1057]. The foreign author had to be present in Washington D.C. on or one day prior to publication to receive copyright in the U.S. Tension between American publishers and European authors remained palpable well into the early 20^th^ century, forcing foreign authors to be strategic about their publication practices to ensures American copyright. Ezra Pound's lament of the "thieving copyright law" as late as the 1920s was representative of European attitudes towards the sole great power not party to the Berne Convention [@Spoo1998, 645]. While the passage of the copyright law is certainly endogenous to the circumstances of the publishing industry [@Safnerpiratethyneighbor], it is beyond the scope of this paper to discuss both the causes and effects of the legal change.\
\
Regardless of the legal imperfections of international copyright, the courtesy of the trade system that had organized "honorable" American piracy of foreign works served little purpose following the 1891 law. While a form of trade courtesy persisted and evolved among American publishers into the mid-20^th^ century, it had little in common with that of its 19^th^ century progenitor. Royalties replaced payments for advance sheets, and @Sheehan1951 [79] finds that foreign authors seem to have obtained about the same terms with American publishers as did American authors. For example, Charles Scribner's Sons royalties on 10 contracts between 1901-1906 seem to have varied between 10-20% for foreign authors and domestic authors alike (ibid). The rule of association, however, was extended to the burgeoning class of American authors, ensuring that publishers would not attempt to poach an author from her associated publisher, unless the author made the first move [@Groves2007b, 141].

[^12]: Also known as the Chace Act.

# Implications

This case study suggests several conclusions about the nature of intellectual property rights. First, it suggests that even in the absence of formal property rights protections on expressions such as written works, it is in the interest of the producers to find ways to protect their property. By treating published works as an commons, always with the tragedy of the commons lurking in the background, the creative industries (in this case, book publishers) share a common interest in not depleting it. Many of the solutions to this problem replicate formal property rights, contracts, and copyright protection by creating exclusivity, the duty to respect others' exclusivity, and a set of remedies for violations. Reciprocity and the mutual recognition of the harms of excessive competition

Second, it suggests the use of the IAD framework to understand the key mechanisms of how a commons can effectively be managed. The courtesy of the trade system, while imperfect, was largely successful between the 1830s and 1891 due to its ability to clearly define boundaries, proportionately link costs and benefits, facilitate collective choices for the industry, enable monitoring of claims, set up graduated sanctions, resolve most conflicts amicably, allow self-organization, and maintain polycentric relations.

Finally, this paper has avoided explicit normative judgment about both the net welfare effect of U.S. piracy of British works and the courtesy of the trade self-regulation within the publishing industry. @Khan2004 --- perhaps the only serious economic study of such effects --- generally finds that consumers benefited from the low prices of books, while U.S. publishers were at least not harmed by lack of foreign copyright protection in the U.S. precisely because of the courtesy of the trade system they set up to simulate copyright. Indeed, Harper wrote in 1877 that:

> The "Law of Trade Courtesy" ... leaves open a way for reprisals on unfair houses, and the people are benefited occasionally by a free fight, in the course of which, while rival publishers are fighting over some tempting morsel, the reading public devours it" [quoted in @Spoo2013, 51].

The system is the interesting combination of several paradoxes. On the one hand, the system allowed a number of American publishers to maintain a large degree of exclusivity over their products to protect profits, but on the other hand, as a voluntary system, the ever-present threat, and occasional intrusion of competition by pirates kept the market contestable, and prices low for consumers. While foreign authors were explicitly denied equal rights on par with their own homeland, or with American authors, the need publishers to establish claims of association led to them compensating foreign authors even when there was no legal duty to do so.

The notion of intellectual property itself remains contentious, with some claiming it is more the result of rent-seeking than a legitimate "right" akin to tangible property [@Boldrin2008; @Kinsella2008; @Bell2014]. Regardless of one's normative view of copyright, it remains in the interest of authors and publishers to establish rules that serve some functions analogous to "legitimate" property rights, whether or not they may be formally recognized.

# References

\singlespacing

### 2021-03-07 manuscript.Rmd
Incorporate Vlad's comments

@@
 ---
-title: Honor Among Thieves^[]
+title: Honor Among Thieves^[I thank Vlad Tarko for helpful comments. All remaining errors are my own.]
 subtitle: How 19^th^ Century American Pirate Publishers Simulated Copyright Protection
@@
 With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce psuedo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection.
-This paper analyzes this system using the Bloomington School's Institutional Analysis and Design (IAD) framework to understand its effectiveness, and pitfalls, in managing the commons of unprotected foreign works in 19^th^ Century America.
+This paper analyzes this system using the Bloomington School's institutional design principles to understand its effectiveness, and pitfalls, in managing the commons of unprotected foreign works in 19^th^ Century America.
   |
@@
 From 1790 to 1891, the United States explicitly prevented foreign authors from obtaining domestic copyright protection for their works.[^1] This policy, in part, was a boon to the domestic publishing industry, which largely reprinted books by foreign authors for an American audience [@Safnerpiratethyneighbor].
-American publishers at the time, deemed "pirates" by their detractors, treated unprotected foreign works as a commons.
-As the United States developed and publishers began to "print on" each other (by reprinting each other's reprints of foreign works), the publishers were forced to develop a system to prevent the foreign-work commons from being depleted from ruinous competition and price wars.
-This system entirely composed of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid to the late 19^th^ century when the U.S.
-began the process of granting copyright protection to foreign authors.
+American publishers at the time, deemed "pirates" by their detractors, treated unprotected foreign works as a "free resource" held in common.
+As American publishers began to "print on" each other (by reprinting each other's reprints of foreign works), they were forced to develop a system to prevent the foreign-work commons from being depleted by ruinous competition and price wars.
+This voluntary system of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid- to the late 19^th^ century, when the U.S.
+began extending copyright protection to foreign authors.
 
@@
 This paper examines the effectiveness of the courtesy of the trade system as a strategy by publishers to manage the common pool resource of unprotected foreign works.
-I use the Bloomington School's IAD framework to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulate the procedures and protections that actual copyright protection brought.
-While it was not uniformly followed by all publishers, saw occasional breakdown into price wars, and ultimately faded with the formal recognition of foreign copyrights, it serves as an interesting example of both how informal norms can resolve disputes, as well as an alternative mechanism of protecting intellectual property without statutory law.
+I use the Bloomington School's "design principles" for robust institutions of governance [@Ostrom1990; @Ostrom2005; @Wilson2013; @Aligica2013; @Tarko2016, Ch.4] to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulate the procedures and protections that actual copyright protection would have legally provided.
+While it was not uniformly followed by all publishers, saw occasional breakdown into price wars, and ultimately faded with the formal recognition of foreign copyrights, trade courtesy serves as an interesting example of both how informal norms can resolve disputes, as well as an alternative mechanism of protecting intellectual property without statutory law.
 
@@
 
-Research by Bloomington School have focused on the importance of informal institutions in managing a commons, cataloging a series of cases and developing a rich grammar of institutions in the Institutional Analysis and Development (IAD) framework [@Ostrom1990; @Hess2007; @McGinnis1999; @Cole2014].
-
-Many researchers focusing broadly on the topic of culture, expression, and knowledge, have productively analyzed these concepts as a commons [@Benkler2002; @Hess2007; @DouradoTabarrok2015; @Schweik2012; @Frischmann2014].
+For 19<sup>th</sup> century U.S.
+publishers, uncopyrighted works by foreign authors constituted a "free" commons, but one quickly subject to tragedy if not properly stewarded.
+Such a nearly "free" resource, however, could easily see its value depleted as other publishers would "print on" another, triggering intense price competition that would evaporate profits.
+The norms and procedures of trade courtesy evolved to precisely prevent this outcome.
+Ironically, many publishers would end up *paying* for this "free" resource via considerable advances to foreign authors to maintain priority in republishing their unprotected works in the U.S.\
+\
+Research by scholars in the Bloomington School tradition have focused on the importance of informal institutions in managing a commons, cataloging a series of cases and developing a rich grammar of institutions and frameworks for evaluating them [@Ostrom1990; @Hess2007; @McGinnis1999; @Cole2014; @Tarko2016].
+In the case of Many researchers focusing broadly on the topic of culture, expression, and knowledge, have productively analyzed these concepts as a commons [@Benkler2002; @Hess2007; @DouradoTabarrok2015; @Schweik2012; @Frischmann2014].
 @Hess2007 [3] take a wider interpretation of commons, defining one as "a resource shared by a group of people that is subject to a social dilemma." [@Ostrom1990,90; @Hess2007,7; @Wilson2013,S22] summarize eight common principles among institutions that successfully manage a commons, paraphrased below:
@@
 
-Such a nearly "free" resource, however, could easily see its value depleted as American publishers began to "print on" other publishers and lowering the price of a particular work in great supply.[^6]
-
-[^6]: Ironically, as described below, as part of the system that emerged to prevent this very outcome, many publishers actually would end up *paying* considerable amounts to foreign authors to maintain priority in republishing their works in America.
-
 The informal "courtesy of the trade" system of norms was both voluntary and ever-changing with market conditions.
@@
 Indeed, the more established publishing houses, which often upheld trade courtesy, saw it as a "gentlemanly" practice, almost the duty of any respectable publisher.
-Holt claimed that none of the major publishers[^7] "would go for another's author any more than for his watch." [quoted in @Spoo2013,40].
+Holt claimed that none of the major publishers[^6] "would go for another's author any more than for his watch." [quoted in @Spoo2013,40].
 
-[^7]: @Spoo2013 lists Putnam, Appleton, Harper, and Scribner fitting this description.
+[^6]: @Spoo2013 lists Putnam, Appleton, Harper, and Scribner fitting this description.
 
@@
 George Putnam, more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," (quoted in @Spoo2013,41).
-By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146].
-\
+By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146].\
 \
@@
 The blood in the water encouraged other publishers to release their own editions of Tennyson, all but evaporating Fields' longstanding claim.
-@Sheehan1951[145] comments, "once a publisher lost control of an association, it could not be reclaimed--- almost as if a copyright had expired and a book had slipped into the public domain."
+@Sheehan1951 [145] comments, "once a publisher lost control of an association, it could not be reclaimed--- almost as if a copyright had expired and a book had slipped into the public domain."
 
@@
 Harper describes how "the contention being commonly left to a fellow-publisher for arbitrament," (quoted in @Spoo2013, 48).
-Correspondance of 40 letters sent by Charles Scribner's Sons to 9 firms[^8] in the 1870s suggest most of these disputes were amicably resolved [@Sheehan1951,70].
+Correspondance of 40 letters sent by Charles Scribner's Sons to 9 firms[^7] in the 1870s suggest most of these disputes were amicably resolved [@Sheehan1951,70].
 
-[^8]: J.B. Lippincott and Co., J.R. Osgood and Co., D. Appleton and Co., Roberts Brothers, G.P. Putnam's Sons, Harper and Brothers, Macmillan and Co., E.P. Dutton and Co., and Henry Holt and Co.
+[^7]: J.B. Lippincott and Co., J.R. Osgood and Co., D. Appleton and Co., Roberts Brothers, G.P. Putnam's Sons, Harper and Brothers, Macmillan and Co., E.P. Dutton and Co., and Henry Holt and Co.
 
@@
 
-Courtesy of the trade was always a voluntary code of norms, slowly internalized over the course of the 19^th^ Century.
-Throughout that period, as described above, various publishers had attempted (and largely failed) to organize trade associations or explicit cartel arrangements, in addition to two periods of fierce price competition between new entrants and established publishers.\
-\
-It was common during the price wars, especially the latter one in the 1870s and 1880s, for upstart firms to enter the market and refuse to abide by trade courtesy [@Spoo2013,47].
-A new class of pirate publishers, like John W.
-Lovell, Isaac K.
-Funk, and George P.
-Munro, emerged with competitive innovations in the 1870s --- creating low quality, cheap "libraries" of reprinted works sent through the mail at subsidized second-class rates [@Groves2007b, 146].
-In 1886 alone, 26 "cheap libraries" issued 1,500 different titles, mostly novels priced between 10-50 cents, thinning the industry's profit margins [@Spoo2013,53].
-Appealing to populist notions of cheap literature for the masses, these entrants directly attacked courtesy as a price-fixing cartel.
-Lovell opined in an 1879 edition of *Publishers' Weekly* that:
-
-> "I can say to the younger and smaller [publishing] houses from my own experience, go in heartily for the 'courtesy of the trade' and---starve.
-You will find everything is expected of you and very little given you.
-As for my part, I prefer to follow the examples that led to success in the past rather than the precepts now advocated to prevent others from attaining it," [quoted in @Groves2007b].
-
-Munro declared that "cheap libraries have broken down the Chinese or rather American wall of trade courtesy and privilege [erected solely for the] monopoly of publishers in this country," [quoted in @Spoo2013, 53-54].
-
-@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^9] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.
-Furthermore, publishing firms, if they were incorporated at all, tended to be closely-held by family outfits.\
-\
-Furthermore, even those firms that upheld trade courtesy knew it was imperfect, and was little more than a paper shield against determined pirates.
-Publishers chief weapon against pirates was speed [@Groves2007b; @Johns2009; @Spoo2013; @Sheehan1951].
-As early as 1823, a letter from Carey in Philadelphia to Sir Walter Scott's Edinburgh publisher boasts:
-
-[^9]: However, this is not necessarily true of certain segments of the book market.
-@Sheehan1951 [50-51] describes the origins of the *American Book Company*, the result of a merger of several leading textbook-publishing firms in the late 1880s, which was estimated to control somewhere between 50-90% of educational textbooks, sufficient to garner antitrust attention by 1909.
-This market, however, is beyond the scope of this paper, which focuses on uncopyrighted foreign works.
-
-> "We have [the] Game completely in our hands this time.
-In 28 hours after receiving [a copy of Scott's*Quentin Durward*], we had 1500 copies off or ready to go...In two days we shall publish it here and in New York, and the Pirates may print it as soon as they please.
-The [pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Spoo2013, 50].
-
-Publishers hired foreign agents in London to obtain intelligence about what firms were to print which British books, and to set up relationships with authors and publishers to safely shuttle works and materials to their press in the U.S.
-before others were able to print.
-Even a handful of hours made all the difference between a profit or a pirate-induced loss on a venture.
-
-## Polycentric external relations
-
+For an institution or set of norms governing a resource to flourish, political authorities must leave sufficient space open for private ordering --- whether by intentional design, or by neglect.
 Trade courtesy in publishing was primarily a response to explicit legal institutions, particularly the gap in copyright protection between domestic and foreign authors.
@@
 (a temporary legal monopoly), but left unaddressed the status of the Stationers' copyright until the 1774 case of *Donaldson v.
-Beckett*, when the House of Lords decided that literary expression is protected only by statutory copyright.[^10] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy of famous works out of London.
+Beckett*, when the House of Lords decided that literary expression is protected only by statutory copyright.[^8] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy of famous works out of London.
 As noted above, many of these tradesmen emigrated to the early United States to carry on their esteemed tradition of pirating English works.
 
-[^10]: The Stationers conjured a clever rhetorical idea that that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the protections of the Statute of Anne, a legal claim the Lords definitely rejected, @Johns2009.
+[^8]: The Stationers conjured a clever rhetorical idea that that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the protections of the Statute of Anne, a legal claim the Lords definitely rejected, @Johns2009.
 
@@
 1239, 1241--42 (C.C.S.D.N.Y.
-1865)[^11] recognized this fact:
+1865)[^9] recognized this fact:
 
-[^11]: Sheldon & Co.
+[^9]: Sheldon & Co.
 and Houghton & Co.
@@
 Interestingly, both parties' attorneys obtained testimony of major publishers about the workings and legitimacy of trade courtesy.
-This is recounted in @Groves2007b[145].
+This is recounted in @Groves2007b [145].
 
@@
 
-The system rapidly broke down after 1891 with the passage of the International Copyright Act.[^12] The Act recognized foreign copyrights in the United States provided that the works themselves were manufactured in the United States.
+The system rapidly broke down after 1891 with the passage of the International Copyright Act.[^10] The Act recognized foreign copyrights in the United States provided that the works themselves were manufactured in the United States.
 This notorious "manufacturing clause," as well as other qualifications and requirements kept the United States ineligible from joining the 1886 Berne Convention until 1989, following major revisions to U.S.
@@
 
-[^12]: Also known as the Chace Act.
+[^10]: Also known as the Chace Act.
+
+## Polycentricity
+
+A major theme of Bloomington School research has been demonstrating the importance of *polycentricity* for robust governance institutions.
+Not merely decentralization, polycentricity implies multiple autonomous decision-makers interacting in a network of shared rules or norms.
+@Aligica2012; @Aligica2013 provide a fleshed-out guide to the logical structure of polycentricity.
+@Tarko2015 [66] summarizes:
+
+> "A polycentric system is a multiplicity of decision centers acting independently but under the constraints of an overarching set of norms and rules which create the conditions for an emergent order to occur via a bottom-up competitive process.
+The key idea is that the overarching set of rules constrains the competitive behavior in the direction of a beneficial emergent outcome."
+
+Often, such polycentric systems feature formal institutions, organizations, or legal/political rules as a major (but crucially, not the only) source of decision-making that affects the community using the shared resource.
+However, as in the case of science [@Tarko2015], courtesy of the trade was entirely a voluntary code of *informal* norms without recourse to formal law (as shown in the previous section), slowly internalized over the course of the 19^th^ Century.
+Throughout that period, as described above, various publishers had attempted (and largely failed) to organize trade associations or explicit cartel arrangements, in addition to two periods of fierce price competition between new entrants and established publishers.
+
+It was common during the price wars, especially the latter one in the 1870s and 1880s, for upstart firms to enter the market and refuse to abide by trade courtesy [@Spoo2013,47].
+A new class of pirate publishers, like John W.
+Lovell, Isaac K.
+Funk, and George P.
+Munro, emerged with competitive innovations in the 1870s --- creating low quality, cheap "libraries" of reprinted works sent through the mail at subsidized second-class rates [@Groves2007b, 146].
+In 1886 alone, 26 "cheap libraries" issued 1,500 different titles, mostly novels priced between 10-50 cents, thinning the industry's profit margins [@Spoo2013,53].
+Appealing to populist notions of cheap literature for the masses, these entrants directly attacked courtesy as a price-fixing cartel.
+Lovell opined in an 1879 edition of *Publishers' Weekly* that:
+
+> "I can say to the younger and smaller [publishing] houses from my own experience, go in heartily for the 'courtesy of the trade' and---starve.
+You will find everything is expected of you and very little given you.
+As for my part, I prefer to follow the examples that led to success in the past rather than the precepts now advocated to prevent others from attaining it," [quoted in @Groves2007b].
+
+Munro declared that "cheap libraries have broken down the Chinese or rather American wall of trade courtesy and privilege [erected solely for the] monopoly of publishers in this country," [quoted in @Spoo2013, 53-54].
+
+@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^11] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.
+Furthermore, publishing firms, if they were incorporated at all, tended to be closely-held by family outfits.\
+\
+Furthermore, even those firms that upheld trade courtesy knew it was imperfect, and was little more than a paper shield against determined pirates.
+Publishers chief weapon against pirates was speed [@Groves2007b; @Johns2009; @Spoo2013; @Sheehan1951].
+As early as 1823, a letter from Carey in Philadelphia to Sir Walter Scott's Edinburgh publisher boasts:
+
+[^11]: However, this is not necessarily true of certain segments of the book market.
+@Sheehan1951 [50-51] describes the origins of the *American Book Company*, the result of a merger of several leading textbook-publishing firms in the late 1880s, which was estimated to control somewhere between 50-90% of educational textbooks, sufficient to garner antitrust attention by 1909.
+This market, however, is beyond the scope of this paper, which focuses on uncopyrighted foreign works.
+
+> "We have [the] Game completely in our hands this time.
+In 28 hours after receiving [a copy of Scott's*Quentin Durward*], we had 1500 copies off or ready to go...In two days we shall publish it here and in New York, and the Pirates may print it as soon as they please.
+The [pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Spoo2013, 50].
+
+Publishers hired foreign agents in London to obtain intelligence about what firms were to print which British books, and to set up relationships with authors and publishers to safely shuttle works and materials to their press in the U.S.
+before others were able to print.
+Even a handful of hours made all the difference between a profit or a pirate-induced loss on a venture.
 
@@
 piracy of British works and the courtesy of the trade self-regulation within the publishing industry.
-@Khan2004 --- perhaps the only serious economic study of such effects --- generally finds that consumers benefited from the low prices of books, while U.S.
+@Khan2004 generally finds that consumers benefited from the low prices of books, while U.S.
 publishers were at least not harmed by lack of foreign copyright protection in the U.S.

### 2021-06-02 manuscript.Rmd
Fix bib

@@
 With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce psuedo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection.
-This paper analyzes this system using the Bloomington School's institutional design principles to understand its effectiveness, and pitfalls, in managing the commons of unprotected foreign works in 19^th^ Century America.
+This paper analyzes this system using the Bloomington School's institutional design principles to understand its effectiveness and pitfalls in managing the commons of unprotected foreign works in 19^th^ Century America.
   |
@@
 
-> "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience...At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Spoo2013 [42]
+> "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience ...
+At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Gilbert1993[31]
 
@@
 As American publishers began to "print on" each other (by reprinting each other's reprints of foreign works), they were forced to develop a system to prevent the foreign-work commons from being depleted by ruinous competition and price wars.
-This voluntary system of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid- to the late 19^th^ century, when the U.S.
+This voluntary system of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid- to the late-19^th^ century, when the U.S.
 began extending copyright protection to foreign authors.
@@
 by U.S.
-printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention standardizing international copyright until 1989.
+printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention, which standardized international copyright, until 1989.
 
@@
 A classic solution, among economists, has been to establish property rights over the resource, allowing owners to exclude other users, internalizing the benefits and responsibility to improve and stewarding over the resource [@Alchian1965; @Demsetz1964].
-However, establishing and enforcing property rights is costly, and are only done when the net benefits of "propertization" exceed the net costs of maintenance (or the net benefits of a commons) [@Demsetz1969].
+However, establishing and enforcing property rights is costly, and is only done when the net benefits of "propertization" exceed the net costs of maintenance (or the net benefits of a commons) [@Demsetz1969].
 Indeed, economists have shown that if rights are designed poorly and allocated from above, then a symmetric tragedy of the *anti*-commons may result, where any one possessor of exclusion-rights among many can veto the productive use of a resource [@Heller1997; @Buchanan2000; @Heller2008].
@@
 Such a nearly "free" resource, however, could easily see its value depleted as other publishers would "print on" another, triggering intense price competition that would evaporate profits.
-The norms and procedures of trade courtesy evolved to precisely prevent this outcome.
+The norms and procedures of trade courtesy evolved precisely to prevent this outcome.
 Ironically, many publishers would end up *paying* for this "free" resource via considerable advances to foreign authors to maintain priority in republishing their unprotected works in the U.S.\
@@
 Research by scholars in the Bloomington School tradition have focused on the importance of informal institutions in managing a commons, cataloging a series of cases and developing a rich grammar of institutions and frameworks for evaluating them [@Ostrom1990; @Hess2007; @McGinnis1999; @Cole2014; @Tarko2016].
-In the case of Many researchers focusing broadly on the topic of culture, expression, and knowledge, have productively analyzed these concepts as a commons [@Benkler2002; @Hess2007; @DouradoTabarrok2015; @Schweik2012; @Frischmann2014].
-@Hess2007 [3] take a wider interpretation of commons, defining one as "a resource shared by a group of people that is subject to a social dilemma." [@Ostrom1990,90; @Hess2007,7; @Wilson2013,S22] summarize eight common principles among institutions that successfully manage a commons, paraphrased below:
+Many researchers focusing broadly on the topic of culture, expression, and knowledge, have productively analyzed these concepts as a commons [@Benkler2002; @Hess2007; @DouradoTabarrok2015; @Schweik2012; @Frischmann2014].
+@Hess2007 [3] take a wider interpretation of a commons, defining one as "a resource shared by a group of people that is subject to a social dilemma." [@Ostrom1990,90; @Hess2007,7; @Wilson2013,S22] summarize eight common principles among institutions that successfully manage a commons, paraphrased below:
 
@@
 @DouradoTabarrok2015 call this the "eightfold path to success" and apply this to ideas and inventions, which they see as a "Super-Lockean commons" which, if accessible, actually *grows* the more that people productively use it (c.f.
-@Hess2007,5).
+@Hess2007 [5]).
 @Safner2016 applies these 8 principles to explain the success of Wikipedia, the free online encyclopedia, in managing public knowledge.
@@
 Many have examined the legal development of American (and previously, British) copyright laws, noting the lapse in protection to foreign authors in the United States throughout the 19^th^ century (among the most systematic are @Patterson1968 and @Johns2009).
-A number have also examined the movement for international copyright in the United States that achieved partial victory at the end of the 1880s (see e.g.
-@Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).
+A number have also examined the movement for international copyright in the United States that achieved partial victory at the end of the 1880s (see, e.g., @Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).
 
-While there is a vast literature in the economics of copyright and intellectual property more broadly,[^2] economists have largely overlooked the dynamics of the event except in the service of broader arguments or economic history regarding intellectual property rights.
+While there is a vast literature in the economics of copyright and intellectual property more broadly,[^2] economists have largely overlooked the dynamics of this period of "piracy" except in the service of broader arguments or economic history regarding intellectual property rights.
 @Khan2005, however, recognizes the idiosyncratic and endogenous component of copyright history, primarily to distinguish the unique and superior American system of IP from that of European systems.
@@
 @Spoo2013 [Ch.1] notably places the workings of the courtesy system in the literature of solutions to the tragedy of the commons, citing touchstone works such as @Ellickson1994 on neighbors solving disputes and @Rustiala2012 on norms in fashion, comedy, and cooking.
-However, the system has yet to be explicitly analyzed in an IAD framework to explain its origin, mechanisms accounting for its success, and decline, as an institution.
+However, the system has yet to be explicitly analyzed in a fuller framework to explain its origin, mechanisms accounting for its success, and decline, as an institution.
 This paper examines how the set of norms and practices constituting "trade courtesy" in the 19<sup>th</sup> century adhere to the 8 principles of success outlined above, as well as how they led to the emergence and eventual decline of this system of self-regulation.
@@
 Thinkers of this school saw the reprinting (authorized or not) of European (predominantly British) works as key to instilling mass literacy, inspiration for innovation, instilling republican virtue [@Johns2009; @Spoo2013, 28; @Clark1960, 70].
-In the first few decades of the American colonies and early republic, the great distances between major cities kept parochial printing presses from cannibalizing each's hold over their local market.
+In the first few decades of the American colonies and early republic, the great distances between major cities kept parochial printing presses from cannibalizing each other's hold over their local market.
 
@@
 The informal "courtesy of the trade" system of norms was both voluntary and ever-changing with market conditions.
-Not all publishers abided by trade courtesy, and some that followed norms closely for one decade would later violate them when it was to their benefit later.
+Not all publishers abided by trade courtesy, and some that followed norms closely for one decade would later violate them when it was to their benefit.
 Some firms that had cut their teeth in the profession by reprinting other publishers' reprints of foreign works later turned "legitimate" by following trade courtesy and enforcing it against upstart pirates.
 
-The objective of trade courtesy was to create a process by which pseudo-property rights could be acquired, transferred, and violations be remedied in a way that would be recognized by enough participants to be reasonably self-enforcing.
-As such, the *raison d'etre* of the system was to define the boundaries over to what did and did not constitute a publisher's "property," and to resolve disputes over those boundaries.
+The objective of the system of trade courtesy was to create a process by which pseudo-property rights could be acquired, transferred, and violations remedied in a way that would be recognized by enough participants to be reasonably self-enforcing.
+As such, the *raison d'etre* of the system was to define the boundaries over what did and did not constitute a publisher's "property," and to resolve disputes over those boundaries.
 
@@
 Publishers following trade courtesy required a firm seeking to stake a claim to undertake a more costly action, often reaching a public agreement with the foreign author, or more frequently, purchasing advance sheets of the edition abroad.
-Holt summarizes: "If a publisher had the advance sheets in his possession, such right or claim overrode a simple announcement," (quoted in @Spoo2013, 37).
-@Spoo2013 [37] further describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown, and the firm undertaking a risky experiment.
+Holt summarizes: "If a publisher had the advance sheets in his possession, such right or claim overrode a simple announcement," (quoted in @Harper1912, 110-11).
+@Spoo2013 [37] describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown, and the firm undertaking a risky experiment.
 
@@
 
-In order for sustainable management of a commons, users must obtain benefits proportionate to the costs they bear.
+To sustainably manage a commons, users must obtain benefits proportionate to the costs they bear.
 Otherwise, if a substantial subset of the group free rides without contributing to the commons, or contributes without any reward, the tragedy of the commons may rear its ugly head, and the resource will be depleted.
-In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs.\
-\
+In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs.
+
 The primary mechanism for upholding the system of norms and preventing tragedy was reciprocity between publishers --- publishers would respect the pseudo-property rights of other publishers to particular works in the hope that others would respect their own claims.
-To reiterate Holt's message, "anybody is welcome who will behave himself."
+To reiterate Holt's message, "anybody is welcome who will behave himself [sic]."
 
@@
 Indeed, the more established publishing houses, which often upheld trade courtesy, saw it as a "gentlemanly" practice, almost the duty of any respectable publisher.
-Holt claimed that none of the major publishers[^6] "would go for another's author any more than for his watch." [quoted in @Spoo2013,40].
+In an Op-ed in *Publisher's Weekly*, Holt claimed that none of the major publishers[^6] "would go for another's author any more than for his watch." [@Publisherslibel,361].
 
@@
 Upholding proper "gentlemanly" conduct was not costless, it did require publishers to sacrifice profitable opportunities in the hope of not destroying their reputation (and hopes of future profit opportunities).
-As mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher $A$ seeking to reprint works of foreign author \$X\$ (and hoping to beat publisher \$B\$ to the punch) could not merely assert this and hope publisher \$B\$ would back down; publisher \$A\$ needs to strike an exclusive arrangement with author \$X\$ or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.
+As mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher *A* seeking to reprint works of foreign author *X* (and hoping to beat publisher *B* to the punch) could not merely assert this and hope publisher *B* would back down; publisher *A* needs to strike an exclusive arrangement with author *X* or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.
 
@@
 Occasionally, it would be payments to the British publisher for advance sheets.
-Harper & Brothers, for example, paid Edward Bulwer-Lytton £50 per volume, Charles Dickens between £250 and £2,000 for his novels [@Spoo2013, 39].
+Harper & Brothers, for example, paid Edward Bulwer-Lytton £50 per volume and Charles Dickens between £250 and £2,000 for his novels [@Spoo2013, 39].
 Carey paid Sir Walter Scott £75 for advance sheets of each of the Waverley novels and £300 for *Life of Napoleon Bonaparte*.
@@
 
-> "In the absence of an international copyright, it is the custom for an English author, or his agent in London, to send early sheets to some American publisher, fixing a price therefor, and by a law of courtesy the American publisher who has issued the previous works of an author is entitled to the first consideration of that author's new book" (quoted in @Spoo2013,39).
+> "In the absence of an international copyright, it is the custom for an English author, or his agent in London, to send early sheets to some American publisher, fixing a price therefor, and by a law of courtesy the American publisher who has issued the previous works of an author is entitled to the first consideration of that author's new book," [@Harper1912,358].
 
@@
 This practice gave an aura of legitimacy to the courtesy system, at least before members of the public (but certainly not authors).
-The *American Publishers' Circular and Literary Gazette* boldly declared in 1872 that it was "an undeniable fact that there is no living English author of established reputation, whose books are extensively republished in this country, who is not freely and properly compensated by the American publisher," (quoted in @Sheehan1951,68).
-George Putnam, more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," (quoted in @Spoo2013,41).
-By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146].\
-\
+The *American Publishers' Circular and Literary Gazette* boldly declared in 1872 that it was "an undeniable fact that there is no living English author of established reputation, whose books are extensively republished in this country, who is not freely and properly compensated by the American publisher," (quoted in @Sheehan1951 [68]).
+George Putnam, more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," [@Putnam1915,365].
+By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146].
+
 Of course, the price of exclusivity on the most popular novels and authors was prohibitively high.
 The temptation to pirates of printing a "sure thing" was too great for any courtesy-abiding publisher to purchase exclusivity with any expectation of it being respected [@Groves2007b, 144].
-Similarly, "classic" (often dead) authors and collected works were unable to be associated with any single publisher.\
-\
+Similarly, "classic" (often dead) authors and collected works were unable to be associated with any single publisher.
+
 In general, the cost of a publisher creating an association with a foreign author under trade courtesy was proportionate to the benefits of expecting others to refrain from printing on one's works.
@@
 Chronicling the history of the corpus of practices, Holt comments: "Trade courtesy is as full of exceptions as the law itself.
-It has grown up as a mass of decisions in particular cases, just as the common law has," (quoted in @Spoo2013, 38).
+It has grown up as a mass of decisions in particular cases, just as the common law has," [@Publisherslibel,360].
 
@@
 Over the century, some publishers occasionally attempted to organize into more formal and explicit associations or cooperate as a large corporation, in order to rationalize the industry.
-Some, like Henry Carey, not-so-secretly dreamed of organizing an American guild like the Stationers Company back in England.
-However, such a scheme always failed because it lacked the flexibility of the courtesy system in two important ways [@Johns2009,197-202]: First, such a corporation would generate a list of titles (akin to the Register of the Stationers Company) available only to its members --- boosting a signal to rogue reprinters that these titles were "sure things" worth printing on.
+Some, like Henry Carey, not-so-secretly dreamed of organizing an American guild like the Stationers Company back in England.[^7] However, such a scheme always failed because it lacked the flexibility of the courtesy system in two important ways [@Johns2009,197-202]: First, such a corporation would generate a list of titles (akin to the Register of the Stationers Company) available only to its members --- boosting a signal to rogue reprinters that these titles were "sure things" worth printing on.
 Second, a formal corporation prevented the very bargains and exchanges between publishers to settle psuedo-property rights disputes to prevent American publishers printing on each other.
+
+[^7]: See section 2.7 for a discussion of the Stationers' Company.
 
@@
 
-Any set of norms or institutions for managing a commons requires diligent monitoring of users by other users, to ensure that no user is free riding and failing in their duties of stewardship.\
-\
+Any set of norms or institutions for managing a commons requires diligent monitoring of users by other users, to ensure that no user is free riding and failing in their duties of stewardship.
+
 While trade courtesy norms could not sustain themselves without vigorous monitoring, each individual publisher had a strong incentive to monitor others for violating norms.
 If a publisher's works were printed on by other publishers, that publisher stood to lose profits from competition.
-Thus, each individual publisher had an interest in ensuring that other publishers respected their claims, and via reciprocity in this manner, the system of trade courtesy was largely self-enforcing.\
-\
-However, it is still instructive to see how publishers commonly took several types of actions in order to notify others of their claims and dispute the claims of others in a public manner.\
-\
+Thus, each individual publisher had an interest in ensuring that other publishers respected their claims, and via reciprocity in this manner, the system of trade courtesy was largely self-enforcing.
+However, it is still instructive to see how publishers commonly took several types of actions in order to notify others of their claims and dispute the claims of others in a public manner.
+
 Since the early days of trade courtesy, the publishers were often "men of letters" and frequently wrote to one another and occasionally published their correspondence for others in their network to see.
@@
 We beg to call your attention to our previous announcement (March 29th) and to an early copy of the work in hand.
-We hope therefore that you will not interfere with us in the matter," quoted in @Sheehan1951,69.
+We hope therefore that you will not interfere with us in the matter," quoted in @Sheehan1951 [69].
 
@@
 @Spoo2013 [43] cites an example in 1870 where the Harpers attempted to print on Holt's edition of Hippolyte Taine's *On Intelligence*.
-Holt wrote to the Harpers merely asking "Doesn't the fact that we have published several of [Taine's] books entitle us to [exclusively publishing \*On Intelligence\*] if we want it?" Such a simple response was sufficient for the Harpers to back off, abiding by the rule of association.
+Holt wrote to the Harpers merely asking "Doesn't the fact that we have published several of [Taine's] books entitle us to [exclusively publishing *On Intelligence*] if we want it?" Such a simple response was sufficient for the Harpers to back off, abiding by the rule of association.
 Upon similarly resolving a second dispute between the firms, Holt graciously expressed that Harper had followed "what the notions of honor...prevalent among publishers of standing required," (ibid).
@@
 Firms that were associated with foreign authors might protect their claim against other publishers printing on them by including copies of letters penned by the foreign author at the beginning of the book.
-Ticknor & Fields editions of Thomas De Quincey's works contained a letter featuring the latter approving of the former "exclusively" publishing his works in America, as did United States Book Co.
-demonstrating they obtained by Rudyard Kipling's "authority," the same for his *Mine Own People* [@Spoo2013,42].
+Ticknor & Fields' editions of Thomas De Quincey's works contained a letter featuring the latter approving of the former "exclusively" publishing his works in America.
+Similarly, United States Book Co.
+showcased Rudyard Kipling's "authority" for them to exclusively publish his *Mine Own People* [@Spoo2013,42].
 
@@
 The ultimate sanction against a "pirate", which if not kept in check by agreement that its *occasional* use was legitimate, was retaliation by predatory pricing.
-While the *raison d'etre* of trade courtesy was to keep prices sufficiently high to avoid a price war, publishers recognized it as legitimate as a response or a preemptive strike against pirates.
-If firm $A$ prints on firm \$B\$ and the latter has a clear claim to priority, firm \$B\$ might reissue the same work below cost to drive firm \$A\$ out of the market (at least for that title).
+While the *raison d'etre* of trade courtesy was to keep prices sufficiently high to avoid a price war, publishers recognized it as a legitimate response or a preemptive strike against pirates.
+If firm *A* prints on firm *B* and the latter has a clear claim to priority, firm *B* might reissue the same work below cost to drive firm *A* out of the market (at least for that title).
 Joseph Henry Harper, of Harper & Brothers wrote:
 
-> "If a publisher declined to comply with the requirements of trade courtesy, some method would be adopted to discipline the offender---generally by the printing of lower-priced editions of his foreign reprints by his aggrieved competitor" (quoted in @Spoo2013,46).
+> "If a publisher declined to comply with the requirements of trade courtesy, some method would be adopted to discipline the offender---generally by the printing of lower-priced editions of his foreign reprints by his aggrieved competitor" [@Harper1912,111-112].
 
-\
-Print on other publishers' claims ran frequent during the two price wars of the 19<sup>th</sup> century.
+Printing on other publishers' claims ran frequent during the two price wars of the 19<sup>th</sup> century.
 In the 1870s, for example, Harper Brothers reprinted their previously published foreign novels at just 10 cents a piece.
@@
 
-> "We determined that [the cheap reprinters] should not share our profits, because we intended that there should be no profit for a division. We began to print on ourselves" [quoted in @Spoo2013,46].
+> "We determined that [the cheap reprinters] should not share our profits, because we intended that there should be no profit for a division. We began to print on ourselves" [@Harper1912,447].
 
-\
 Upstart firms were less able to withstand below-cost prices than established publishing houses.
@@
 Similarly, Ticknor & Fields and Harper found themselves in a dispute over issuing Charles Dickens' *Mystery of Edwin Drood*, ultimately agreeing that Ticknor & Fields would issue the novel as a single book and Harper would bring it out serially in *Harper's Weekly* [@Spoo2013,48].
-A common solution was for one claimaint to issue an edition in (relatively more expensive) cloth, and another claimant to issue the same work in cheaper paperback; alternatively, if the work were one of a series, one claimant release the disputed volume, and another release the following volume, @Sheehan1951 [74].
+A common solution was for one claimaint to issue an edition in (relatively more expensive) cloth, and another claimant to issue the same work in cheaper paperback; alternatively, if the work were one of a series, one claimant released the disputed volume, and another released the following volume, [@Sheehan1951,74].
 Much like the owner of a true copyright today, a publisher with a good claim to issuing a foreign work in the U.S.
 could bargain with other publishers over a wide range of in-kind rights - issuing serials, hardbound books, paperback books, etc.
-- to reach an amicable arrangement.\
-\
+- to reach an amicable arrangement.
+
 Publishers that were unable to reach bilateral agreement often submitted the competing claims before an agreed upon third publisher to act as a neutral arbitrator.
-Harper describes how "the contention being commonly left to a fellow-publisher for arbitrament," (quoted in @Spoo2013, 48).
-Correspondance of 40 letters sent by Charles Scribner's Sons to 9 firms[^7] in the 1870s suggest most of these disputes were amicably resolved [@Sheehan1951,70].
+Harper describes how "the contention being commonly left to a fellow-publisher for arbitrament," [@Harper1912, 111-112].
+Correspondence of 40 letters sent by Charles Scribner's Sons to 9 firms[^8] in the 1870s suggest most of these disputes were amicably resolved [@Sheehan1951,70].
 
-[^7]: J.B. Lippincott and Co., J.R. Osgood and Co., D. Appleton and Co., Roberts Brothers, G.P. Putnam's Sons, Harper and Brothers, Macmillan and Co., E.P. Dutton and Co., and Henry Holt and Co.
+[^8]: J.B. Lippincott and Co., J.R. Osgood and Co., D. Appleton and Co., Roberts Brothers, G.P. Putnam's Sons, Harper and Brothers, Macmillan and Co., E.P. Dutton and Co., and Henry Holt and Co.
 
@@
 Like most medieval craft guilds, the English crown granted the Stationers the sole right to print within the realm, and for them to govern their own trade.
-While often politically controversial in the tumultuous English Civil Wars of the 17^th^ Century [@Safnerpatronage], the main feature of the Stationers' practices was the establishment of a Register that maintained whom had the exclusive, perpetual, and transferable right to print a particular work (known as "a copy right") as well as a series of sanctions and methods of adjudicating disputes based on copies authorized by the Register.
+While often politically controversial in the tumultuous English Civil Wars of the 17^th^ Century [@Safnerpatronage], the main feature of the Stationers' practices was the establishment of a Register that maintained who had the exclusive, perpetual, and transferable right to print a particular work (known as "a copy right") as well as a series of sanctions and methods of adjudicating disputes based on copies authorized by the Register.
 Alongside this grew up "trade customs" of printing as an honorable craft.
@@
 (a temporary legal monopoly), but left unaddressed the status of the Stationers' copyright until the 1774 case of *Donaldson v.
-Beckett*, when the House of Lords decided that literary expression is protected only by statutory copyright.[^8] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy of famous works out of London.
+Beckett*, when the House of Lords decided that literary expression is protected only by statutory copyright.[^9] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy of famous works out of London.
 As noted above, many of these tradesmen emigrated to the early United States to carry on their esteemed tradition of pirating English works.
 
-[^8]: The Stationers conjured a clever rhetorical idea that that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the protections of the Statute of Anne, a legal claim the Lords definitely rejected, @Johns2009.
+[^9]: The Stationers conjured a clever rhetorical idea that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the protections of the Statute of Anne, a legal claim the Lords definitely rejected, @Johns2009.
 
@@
 United States*, 1945 (No.
-4491) at 20; quoted in @Spoo1998,70).
+4491) at 20; quoted in @Spoo1998 [70]).
 
@@
 1239, 1241--42 (C.C.S.D.N.Y.
-1865)[^9] recognized this fact:
+1865)[^10] recognized this fact:
 
-[^9]: Sheldon & Co.
+[^10]: Sheldon & Co.
 and Houghton & Co.
@@
 > [trade courtesy] confessedly rests upon no common law of the country, recognized and administered by judicial tribunals.
-If it has any foundation at all, it stands on the mere will, or...the "courtesy" of the trade...It can, therefore, hardly be called property at all---certainly not in any sense known to the law, (quoted in @Spoo2013, 34).
+If it has any foundation at all, it stands on the mere will, or...the "courtesy" of the trade...It can, therefore, hardly be called property at all---certainly not in any sense known to the law, quoted in [@Spoo2013, 34].
 
-The system rapidly broke down after 1891 with the passage of the International Copyright Act.[^10] The Act recognized foreign copyrights in the United States provided that the works themselves were manufactured in the United States.
-This notorious "manufacturing clause," as well as other qualifications and requirements kept the United States ineligible from joining the 1886 Berne Convention until 1989, following major revisions to U.S.
+The system rapidly broke down after 1891 with the passage of the International Copyright Act.[^11] The Act recognized foreign copyrights in the United States provided that the works themselves were manufactured in the United States.
+This notorious "manufacturing clause," as well as other qualifications and requirements kept the United States ineligible to join the 1886 Berne Convention until 1989, following major revisions to U.S.
 law.
@@
 Register of Copyrights Barbara Ringer comments that the law had enough loopholes to "make the extension of copyright protection to foreigners illusory" [@Ringer1968,1057].
-The foreign author had to be present in Washington D.C.
+The foreign author had to be present in Washington, D.C.
 on or one day prior to publication to receive copyright in the U.S.
-Tension between American publishers and European authors remained palpable well into the early 20^th^ century, forcing foreign authors to be strategic about their publication practices to ensures American copyright.
+Tension between American publishers and European authors remained palpable well into the early 20^th^ century, forcing foreign authors to be strategic about their publication practices to ensure American copyright.
 Ezra Pound's lament of the "thieving copyright law" as late as the 1920s was representative of European attitudes towards the sole great power not party to the Berne Convention [@Spoo1998, 645].
@@
 Royalties replaced payments for advance sheets, and @Sheehan1951 [79] finds that foreign authors seem to have obtained about the same terms with American publishers as did American authors.
-For example, Charles Scribner's Sons royalties on 10 contracts between 1901-1906 seem to have varied between 10-20% for foreign authors and domestic authors alike (ibid).
+For example, Charles Scribner's Sons' royalties on 10 contracts between 1901-1906 seem to have varied between 10-20% for foreign authors and domestic authors alike (ibid).
 The rule of association, however, was extended to the burgeoning class of American authors, ensuring that publishers would not attempt to poach an author from her associated publisher, unless the author made the first move [@Groves2007b, 141].
 
-[^10]: Also known as the Chace Act.
+[^11]: Also known as the Chace Act.
 
@@
 Often, such polycentric systems feature formal institutions, organizations, or legal/political rules as a major (but crucially, not the only) source of decision-making that affects the community using the shared resource.
-However, as in the case of science [@Tarko2015], courtesy of the trade was entirely a voluntary code of *informal* norms without recourse to formal law (as shown in the previous section), slowly internalized over the course of the 19^th^ Century.
+However, as in the case of science [@Tarko2015], courtesy of the trade was entirely a voluntary code of *informal* norms without recourse to formal law (as shown in the previous section), slowly internalized over the course of the 19^th^ century.
 Throughout that period, as described above, various publishers had attempted (and largely failed) to organize trade associations or explicit cartel arrangements, in addition to two periods of fierce price competition between new entrants and established publishers.
@@
 You will find everything is expected of you and very little given you.
-As for my part, I prefer to follow the examples that led to success in the past rather than the precepts now advocated to prevent others from attaining it," [quoted in @Groves2007b].
+As for my part, I prefer to follow the examples that led to success in the past rather than the precepts now advocated to prevent others from attaining it," quoted in [@Groves2007b].
 
@@
 
-@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^11] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.
+@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^12] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.
 Furthermore, publishing firms, if they were incorporated at all, tended to be closely-held by family outfits.\
@@
 Furthermore, even those firms that upheld trade courtesy knew it was imperfect, and was little more than a paper shield against determined pirates.
-Publishers chief weapon against pirates was speed [@Groves2007b; @Johns2009; @Spoo2013; @Sheehan1951].
+Publishers' chief weapon against pirates was speed [@Groves2007b; @Johns2009; @Spoo2013; @Sheehan1951].
 As early as 1823, a letter from Carey in Philadelphia to Sir Walter Scott's Edinburgh publisher boasts:
 
-[^11]: However, this is not necessarily true of certain segments of the book market.
+[^12]: However, this is not necessarily true of certain segments of the book market.
 @Sheehan1951 [50-51] describes the origins of the *American Book Company*, the result of a merger of several leading textbook-publishing firms in the late 1880s, which was estimated to control somewhere between 50-90% of educational textbooks, sufficient to garner antitrust attention by 1909.
@@
 > "We have [the] Game completely in our hands this time.
-In 28 hours after receiving [a copy of Scott's*Quentin Durward*], we had 1500 copies off or ready to go...In two days we shall publish it here and in New York, and the Pirates may print it as soon as they please.
-The [pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Spoo2013, 50].
+In 28 hours after receiving [a copy of Scott's *Quentin Durward*], we had 1500 copies off or ready to go...In two days we shall publish it here and in New York, and the Pirates may print it as soon as they please.
+The [pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Madison1966, 10].
 
@@
 First, it suggests that even in the absence of formal property rights protections on expressions such as written works, it is in the interest of the producers to find ways to protect their property.
-By treating published works as an commons, always with the tragedy of the commons lurking in the background, the creative industries (in this case, book publishers) share a common interest in not depleting it.
+By treating published works as a commons, always with the tragedy of the commons lurking in the background, the creative industries (in this case, book publishers) share a common interest in not depleting it.
 Many of the solutions to this problem replicate formal property rights, contracts, and copyright protection by creating exclusivity, the duty to respect others' exclusivity, and a set of remedies for violations.
-Reciprocity and the mutual recognition of the harms of excessive competition
+Reciprocity and the mutual recognition of the harms of excessive competition were sufficient for most publishers to behave *as if* individual publishers had exclusive property rights over their foreign reprints.
 
@@
 > The "Law of Trade Courtesy" ...
-leaves open a way for reprisals on unfair houses, and the people are benefited occasionally by a free fight, in the course of which, while rival publishers are fighting over some tempting morsel, the reading public devours it" [quoted in @Spoo2013, 51].
+leaves open a way for reprisals on unfair houses, and the people are benefited occasionally by a free fight, in the course of which, while rival publishers are fighting over some tempting morsel, the reading public devours it" [@Harper1912, 393].
 

### 2021-06-08 manuscript.Rmd
Prepped for journal

@@
 ---
-title: Honor Among Thieves^[I thank Vlad Tarko for helpful comments. All remaining errors are my own.]
+title: Honor Among Thieves^[I thank Vlad Tarko and Elana Safner for helpful comments. All remaining errors are my own.]
 subtitle: How 19^th^ Century American Pirate Publishers Simulated Copyright Protection
@@
   | *JEL Classification:* H41, K29, N40, O34
-  | *Keywords:* Copyright, intellectual property, economic history
+  | *Keywords:* Copyright, intellectual property, literary piracy, publishing, economic history
 
@@
 > "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience ...
-At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Gilbert1993[31]
+At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Gilbert1993 [31]
 
@@
 The norms and procedures of trade courtesy evolved precisely to prevent this outcome.
-Ironically, many publishers would end up *paying* for this "free" resource via considerable advances to foreign authors to maintain priority in republishing their unprotected works in the U.S.\
+Ironically, publishers typically ended up *paying* for this "free" resource by fronting considerable advances to foreign authors to maintain priority in republishing their unprotected works in the U.S.\
 \
@@
 
-At mid-century, there were about 400 publishers, with a supermajority of them in New York, Philadelphia, Baltimore, and Boston, according to the authoritative *Trubner's Bibliographical Guide to American Literature*, compiled in 1859, [@Shove1936, 23].
+At mid-century, there were about 400 publishers concentrated in New York, Philadelphia, Baltimore, and Boston, according to the authoritative *Trubner's Bibliographical Guide to American Literature*, compiled in 1859, [@Shove1936, 23].
 Harper & Brothers was the largest firm, with over 2 million volumes sold in 1859.
 
-The growth of the industry brought about fierce competition, especially after the Depression in the late 1830s and a ruinous price war into the 1840s.
-The publisher Carey & Lea, for example, had seen their output of new titles shrink by over 50% between 1835 and 1837 [@Groves2007b, 141-142].
-Publishers also resorted to clever alternative forms of competition, such as the "weeklies" which printed entire English novels on "mammoth" size single sheets sent through the mail direct to customers, akin to newspapers [@Groves2007b, 142; @Barnes1974, 4-24].
-These changes underscored the need for self-regulation and organizational solutions within the industry into the 1840s.
+The growth of the industry brought about fierce competition, especially after the Depression in the late 1830s and a ruinous price war in the 1840s.
+The Philadelphia publisher Carey & Lea, for example, saw their output of new titles shrink by over 50% between 1835 and 1837 [@Groves2007b, 141-142].
+Upstart periodicals such as *Brother Jonathan* and *New World* responded to an uptick in demand for literature by resorting to clever alternative forms of publishing: "monthlies" and "weeklies" printed entire English novels in small, low quality type on "mammoth" size single sheets sent through the mail direct to customers in newspaper format, exploiting cheap postal rates [@Groves2007b, 142; @Barnes1974, 4-24; @Madison1966,24-25].
+This heavily depressed book prices, forcing publishers to respond in kind (as did Harper & Brothers), or to exit the market (like Wiley and Blanchard & Lea, both of whom focused on copyrightable textbooks) [@Madison1966,25][^5].
+This period of "cutthroat competition" caused publishers to recognize the need for self-regulation and organizational solutions within the industry into the 1840s.
 Some publishers, such as Carey, suggested forming an explicit organization modeled off of the London Stationers Company.
-Carey set up The Philadelphia Company as an explicit cartel arrangement, only to see it evaporate 5 years later.[^5] Despite its failure, @Johns2009 [199] summarizes the common sentiment within the publishing industry:
+Carey set up The Philadelphia Company as an explicit cartel arrangement, only to see it evaporate 5 years later.[^6] Despite its failure, @Johns2009 [199] summarizes the common sentiment within the publishing industry:
 
-[^5]: Once such a cartel allocated titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint.
+[^5]: For more on the dynamics of the price wars and a general framework for how publishers strategies evolved over the 19th century, see @Safnerpiratethyneighbor.
+
+[^6]: Once such a cartel allocated titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint.
 Without an inflexible cartel, a bargain could be reached between two publishers to establish boundaries, but this was not possible with a large organization and rogue outside printers.
@@
 Indeed, the more established publishing houses, which often upheld trade courtesy, saw it as a "gentlemanly" practice, almost the duty of any respectable publisher.
-In an Op-ed in *Publisher's Weekly*, Holt claimed that none of the major publishers[^6] "would go for another's author any more than for his watch." [@Publisherslibel,361].
-
-[^6]: @Spoo2013 lists Putnam, Appleton, Harper, and Scribner fitting this description.
+In an Op-ed in *Publisher's Weekly*, Holt claimed that the major publishers, "Putnam, Appleton, Harper and Scribner...were much more inclined to friendly cooperation and mutual concession than barbarous competition...No one of them, or of a few more, would go for another's author any more than for his watch." [@Publisherslibel,361].
 
@@
 This required investment led to one of the great ironies of the system: despite the absence of a legal duty, U.S.
-publishers frequently paid significant honoraria to British authors for the "right" to print their works.
+publishers frequently paid significant honoraria to British authors for the priority "right" to print their works.
 In fact, publishers often *emphasized* in their advertisements the amount they paid to foreigners for their "exclusive" or "authorized" edition [@Spoo2013, 24].
@@
 Harper & Brothers, for example, paid Edward Bulwer-Lytton £50 per volume and Charles Dickens between £250 and £2,000 for his novels [@Spoo2013, 39].
-Carey paid Sir Walter Scott £75 for advance sheets of each of the Waverley novels and £300 for *Life of Napoleon Bonaparte*.
-The Harpers described the courtesy practice of purchasing advance sheets from foreign authors or their agents or publishers:
+Harper Brothers, the largest 19<sup> th</sup> century publisher in the U.S.
+(and later the world) described the courtesy practice of purchasing advance sheets from foreign authors or their agents or publishers:
 
@@
 
-Henry Holt, in his 1893 reflections on trade courtesy, notes that "It not only prevented ruinous competition between American publishers, but also secured to foreign authors most of their rights," (ibid, 32).
-This practice gave an aura of legitimacy to the courtesy system, at least before members of the public (but certainly not authors).
-The *American Publishers' Circular and Literary Gazette* boldly declared in 1872 that it was "an undeniable fact that there is no living English author of established reputation, whose books are extensively republished in this country, who is not freely and properly compensated by the American publisher," (quoted in @Sheehan1951 [68]).
-George Putnam, more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," [@Putnam1915,365].
+This practice of paying for advance sheets gave an aura of semi-legitimacy to the courtesy system.
+Publisher Henry Holt, in his 1893 reflections on trade courtesy, notes that "It not only prevented ruinous competition between American publishers, but also secured to foreign authors most of their rights," (ibid, 32).
+While many popular British authors publicly lamented the lack of copyright protection in America --- notably Charles Dickens, who made it a feature of his 1867 tour of the U.S.
+--- they were often compensated even though publishers had no obligation to do so.
+Such honoraria and payment for the "exclusivity" that came with being first, or with the author's "authorized" version, was surely less than the author would have received in royalties or profit-sharing under copyright protection.
+British author Frederick Marryat estimated in 1839 that if Walter Scott (the British author who first became a best seller in the U.S.) lost out on \$500,000 in royalties from U.S.
+sales [@Madison1966,8-12].
+Carey & Lea, his "authorized" American publisher made "token" payments (by comparison) --- £50 for Canongate in 1824, £250 for Life of Napoleon in 1827 (which was the largest such U.S.
+payment to that date) [@Madison1966,12].
+To further underscore the copyright tradeoff between U.S.
+and Britain, Scott's novels often sold for 1/4 of the British price in the U.S., but gave him more American readers than English readers in total (ibid,12).
+
+Some established publishing houses, such as Putnam and Ticknor & Fields, took it as a point of honor to ensure British authors were properly compensated [@Madison1966,63].
+The *American Publishers' Circular and Literary Gazette* was so bold as to declare in 1872 that it was "an undeniable fact that there is no living English author of established reputation, whose books are extensively republished in this country, who is not freely and properly compensated by the American publisher," (quoted in @Sheehan1951 [68]).
+Publisher George Putnam more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," [@Putnam1915,365].
 By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146].
@@
 Ezra Pound's lament of the "thieving copyright law" as late as the 1920s was representative of European attitudes towards the sole great power not party to the Berne Convention [@Spoo1998, 645].
-While the passage of the copyright law is certainly endogenous to the circumstances of the publishing industry [@Safnerpiratethyneighbor], it is beyond the scope of this paper to discuss both the causes and effects of the legal change.\
+While the passage of the copyright law is certainly endogenous to the market conditions facing the publishing industry [@Safnerpiratethyneighbor], it is beyond the scope of this paper to discuss both the causes and effects of the legal change.\
 \
@@
 
-Munro declared that "cheap libraries have broken down the Chinese or rather American wall of trade courtesy and privilege [erected solely for the] monopoly of publishers in this country," [quoted in @Spoo2013, 53-54].
+Munro declared that "My contemporaries have called me a pirate.
+Posterity will have a truer word with which to characterize my work---that of reformer.
+The cheap libraries have broken down the Chinese or rather American wall of trade courtesy and privilege [erected solely for the] monopoly of publishers in this country," [quoted in @Madison1966, 53].
 
-@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^12] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.
-Furthermore, publishing firms, if they were incorporated at all, tended to be closely-held by family outfits.\
+@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^12] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.\
 \
@@
 Publishers' chief weapon against pirates was speed [@Groves2007b; @Johns2009; @Spoo2013; @Sheehan1951].
-As early as 1823, a letter from Carey in Philadelphia to Sir Walter Scott's Edinburgh publisher boasts:
+As early as 1823, a letter from Mathew Carey in Philadelphia to Sir Walter Scott's Edinburgh publisher boasts:
 
@@
 In 28 hours after receiving [a copy of Scott's *Quentin Durward*], we had 1500 copies off or ready to go...In two days we shall publish it here and in New York, and the Pirates may print it as soon as they please.
-The [pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Madison1966, 10].
+The [rival pirates'] Edition will be out in about 48 hours after they have one of our Copies but we shall have complete and entire possession of every market in the Country for a short time," [quoted in @Madison1966, 10].
 
@@
 before others were able to print.
-Even a handful of hours made all the difference between a profit or a pirate-induced loss on a venture.
+Even a few days or hours made all the difference between a profit or a pirate-induced loss on a venture.
+Harper Brothers, for example, could frequently outrun its largest competitor, Carey & Lea shipping from New York into the Ohio valley, and at half the cost, compared with Philadelphia @Madison1966, 12,21].
 
@@
 
-Second, it suggests the use of the IAD framework to understand the key mechanisms of how a commons can effectively be managed.
+Second, it suggests how using institutional design principles can help us understand the key mechanisms of how a commons can effectively be managed.
 The courtesy of the trade system, while imperfect, was largely successful between the 1830s and 1891 due to its ability to clearly define boundaries, proportionately link costs and benefits, facilitate collective choices for the industry, enable monitoring of claims, set up graduated sanctions, resolve most conflicts amicably, allow self-organization, and maintain polycentric relations.
@@
 The notion of intellectual property itself remains contentious, with some claiming it is more the result of rent-seeking than a legitimate "right" akin to tangible property [@Boldrin2008; @Kinsella2008; @Bell2014].
-Regardless of one's normative view of copyright, it remains in the interest of authors and publishers to establish rules that serve some functions analogous to "legitimate" property rights, whether or not they may be formally recognized.
+Regardless of one's normative view of copyright, it remains in the interest of authors and especially publishers to establish property rights over printed works, whether formal or informal.
 

### 2021-07-30 JLEO_cover.Rmd
Manuscript had typo in abstract!
Also cover letter

---
author:
- Ryan Safner
- Assistant Professor of Economics
- Hood College

opening: Dear Dr. Prat,
closing: Sincerely,
address:
- Dr. Andrea Prat
- Editor-in-Chief
- Journal of Law, Economics, & Organization

signature: "signaturesmall.png" # doesn't like underscores
signature-before: -8ex
signature-after: 0ex

inline-links: true # instead of footnotes

letterhead: "Hoodhead.png"
#letterhead-scale: 0.4
#letterfoot-scale: 0.4
letterfoot: "Hoodfoot.png"
colorlinks: true

output: linl::linl
---

I am writing to submit my manuscript, "Honor Among Thieves: How 19^th^ Century American Pirate Publishers Simulated Copyright Protection." This paper shows a case study of how firms used a gap in copyright law to create their own voluntary institutions to manage a common resource — uncopyrighted foreign works in the 19th century United States.

Many economists tend to consider books, ideas, and other material traditionally protected by intellectual property rights as public goods due to the nonrival nature of information. For 19th Century American publishers of foreign (mostly British) works (which the Copyright Act of 1790 prevented from obtaining copyright), it is better understood as a problem of a depletable commons: while it was legal for any US publisher to print an unauthorized copy of an unprotected foreign work, multiple US publishers could easily "print on" each other to make a quick buck off the same popular work. This led to several ruinous price wars within the industry, and prompted U.S. publishers to create the "courtesy of the trade" system of voluntary norms to regulate printing foreign uncopyrighted works. The primary purpose and effect was to create pseudo-property rights that publishers could establish and assert over individual works, preventing others from "printing on" them and thus staving off price competition. I employ the Bloomington School's institutional design principles to better understand how (well) this system worked, why and when it arose and declined.

I believe this paper contributes to several literatures on copyright and (formal vs. informal) legal institutions broadly by illuminating a case study in the economic history of copyright law and by using the Bloomington's approach to understanding commons. I hope this will contribute to our understanding of the origins, challenges, and potential reforms to modern copyright law. Thank you for the opportunity to submit to your journal.

### 2021-07-30 manuscript.Rmd
Manuscript had typo in abstract!
Also cover letter

@@
   | From 1790 to 1891, the United States prevented foreign authors from obtaining domestic copyright protection, implicitly subsidizing a domestic reprinting industry.
-With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce psuedo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection.
+With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce pseudo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection.
 This paper analyzes this system using the Bloomington School's institutional design principles to understand its effectiveness and pitfalls in managing the commons of unprotected foreign works in 19^th^ Century America.

### 2021-08-11 ALER_cover.Rmd
ALER, and add "commons" to keywords

--- 2021-07-30 JLEO_cover.Rmd
@@
 
-opening: Dear Dr. Prat,
+opening: Dear Dr. Crocco,
 closing: Sincerely,
 address:
-- Dr. Andrea Prat
-- Editor-in-Chief
-- Journal of Law, Economics, & Organization
+- Dr. Karen Crocco
+- Managing Editor
+- American Law and Economics Review
 
@@
 
-I am writing to submit my manuscript, "Honor Among Thieves: How 19^th^ Century American Pirate Publishers Simulated Copyright Protection." This paper shows a case study of how firms used a gap in copyright law to create their own voluntary institutions to manage a common resource — uncopyrighted foreign works in the 19th century United States.
+I am writing to submit my manuscript, "Honor Among Thieves: How 19^th^ Century American Pirate Publishers Simulated Copyright Protection" to your journal.
+This paper shows a case study of how firms used a gap in copyright law to create their own voluntary institutions to manage a common resource — uncopyrighted foreign works in the 19th century United States.
 

### 2022-12-22 response_to_reviewers.qmd
Edits for editor

---
title: "Response to Reviewers"
format: pdf
editor: visual
---

Dear Reviewers,\
\
Thank you very much for reviewing and providing comments on my manuscript, they have been very helpful in improving it. In addition to grammatical fixes and minor edits to improve clarity, I have made two major changes to the paper in response to your suggestions.\
\
First, I have added a short discussion at the end of section 2.1 to respond to Reviewer 1's request for examples of uncommitted publishers. It seems to have been that priority rights were respected without costly commitments (paying for advance sheets, negotiating permission from the author, etc) in cases of new & unknown authors. Here, exclusivity was not seen as too valuable as there was little competition to publish these authors. These are primarily for publishers to undertake experiments, and only if the gamble pays off and the author becomes popular enough to keep reprinting, the firm that first published them still has rights to exclusivity under trade courtesy.

Second, I have added a discussion to expand upon the role of reputation and repeated dealings constraining defection, using the example of a simple repeated Prisoners' Dilemma and trigger strategies, as suggested by the editor. I discuss a firm's reputation as upholding trade courtesy and avoiding piracy as a depreciable capital asset that a publisher risks devaluing if they turn to piracy.

In response to Reviewer 1's question about lesser sanctions (examples of compensation) that prevented piracy - I do provide a few examples of how disputes were settled among followers of courtesy. It is certainly difficult to deter a dedicated pirate publisher with anything but a price war. But for the "legitimate" publishers that occasionally stepped on each other's toes, I provide several examples in that section, as well as section 2.6, about how publishers resolved such disputes with "adjustments" or in-kind bargains where the "rights" are often shared among publishers in different forms - the right to paperback editions vs. hardback, serials, present vs. future editions. Admittedly some of this was a structural decision about where to put these examples in the paper to have sufficient material for each of the 8 design principles.\
\
Thank you again very much for your comments and for the opportunity to submit this paper.

### 2022-12-22 manuscript.Rmd
ALER, and add "commons" to keywords
Edits for editor

@@
   | *JEL Classification:* H41, K29, N40, O34
-  | *Keywords:* Copyright, intellectual property, literary piracy, publishing, economic history
+  | *Keywords:* Copyright, intellectual property, literary piracy, publishing, economic history, commons
 
@@
 > "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience ...
-At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Gilbert1993 [31]
+At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." — Henry Holt, quoted in @Gilbert1993 [31]
 
 From 1790 to 1891, the United States explicitly prevented foreign authors from obtaining domestic copyright protection for their works.[^1] This policy, in part, was a boon to the domestic publishing industry, which largely reprinted books by foreign authors for an American audience [@Safnerpiratethyneighbor].
-American publishers at the time, deemed "pirates" by their detractors, treated unprotected foreign works as a "free resource" held in common.
-As American publishers began to "print on" each other (by reprinting each other's reprints of foreign works), they were forced to develop a system to prevent the foreign-work commons from being depleted by ruinous competition and price wars.
+American publishers at the time — deemed "pirates" by their detractors — treated unprotected foreign works as a "free resource" held in common.
+As American publishers began to "print on" each other (the industry's term for reprinting each other's reprints of foreign works without permission), they were forced to develop a system to prevent the foreign-work commons from being depleted by ruinous competition and price wars.
 This voluntary system of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid- to the late-19^th^ century, when the U.S.
@@
 
-[^1]: Even after the 1891 Chace Act that first gave copyright protection in the U.S.
+[^1]: Even after the 1891 International Copyright Act first gave copyright protection in the U.S.
 to foreign authors, it was *de facto* quite costly for authors to attain this.
@@
 by U.S.
-printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention, which standardized international copyright, until 1989.
+printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention, which standardized international copyright recognition, until 1989.
 
 This paper examines the effectiveness of the courtesy of the trade system as a strategy by publishers to manage the common pool resource of unprotected foreign works.
-I use the Bloomington School's "design principles" for robust institutions of governance [@Ostrom1990; @Ostrom2005; @Wilson2013; @Aligica2013; @Tarko2016, Ch.4] to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulate the procedures and protections that actual copyright protection would have legally provided.
+I use the Bloomington School's "design principles" for robust institutions of governance [@Ostrom1990; @Ostrom2005; @Wilson2013; @Aligica2013; @Tarko2016, Ch.4] to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulated the procedures and protections that actual copyright protection would have legally provided.
 While it was not uniformly followed by all publishers, saw occasional breakdown into price wars, and ultimately faded with the formal recognition of foreign copyrights, trade courtesy serves as an interesting example of both how informal norms can resolve disputes, as well as an alternative mechanism of protecting intellectual property without statutory law.
@@
 Economists typically define a commons, or a common pool resource, as being rivalrous but non-excludable: each person's consumption subtracts from another's ability to consume, but all such users are unable to prevent others from accessing the resource.
-@Hardin1968 famously termed the situation a "tragedy of the commons," as the inability of users to exclude others' use logically implies users will rapidly deplete the resource.
-A classic solution, among economists, has been to establish property rights over the resource, allowing owners to exclude other users, internalizing the benefits and responsibility to improve and stewarding over the resource [@Alchian1965; @Demsetz1964].
-However, establishing and enforcing property rights is costly, and is only done when the net benefits of "propertization" exceed the net costs of maintenance (or the net benefits of a commons) [@Demsetz1969].
+@Hardin1968 famously termed the situation a "tragedy of the commons," as the inability of users to exclude others' use logically implies users will have an individual incentive to overuse the resource, leading to its depletion.
+A classic solution suggested by economists is to establish property rights over the resource which allows owners to exclude other users, internalizing the benefits and responsibility to improve and stewarding over the resource [@Alchian1965; @Demsetz1964].
+However, establishing and enforcing property rights is costly itself, and is only done when the net benefits of "propertization" exceed the net costs of establishing and enforcing property (or the net benefits of a commons) [@Demsetz1969].
 Indeed, economists have shown that if rights are designed poorly and allocated from above, then a symmetric tragedy of the *anti*-commons may result, where any one possessor of exclusion-rights among many can veto the productive use of a resource [@Heller1997; @Buchanan2000; @Heller2008].
@@
 @Hess2007 [5]).
-@Safner2016 applies these 8 principles to explain the success of Wikipedia, the free online encyclopedia, in managing public knowledge.
+@Safner2016wiki applies these 8 principles to explain the success of Wikipedia, the free online encyclopedia, in managing public knowledge.
 
-Federally-sponsored American literary piracy has been noted and extensively discussed by lawyers, historians, and literary scholars.
-Many have examined the legal development of American (and previously, British) copyright laws, noting the lapse in protection to foreign authors in the United States throughout the 19^th^ century (among the most systematic are @Patterson1968 and @Johns2009).
-A number have also examined the movement for international copyright in the United States that achieved partial victory at the end of the 1880s (see, e.g., @Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).
+State-sponsored American literary piracy has been noted and extensively discussed by lawyers, historians, and literary scholars.
+Many researchers have examined the legal development of American (and previously, British) copyright laws, noting the lapse in protection to foreign authors in the United States throughout the 19^th^ century (among the most systematic are @Patterson1968 and @Johns2009).
+A number have also examined the movement for international copyright recognition in the United States that achieved partial victory at the end of the 1880s (see, e.g., @Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).
 
 While there is a vast literature in the economics of copyright and intellectual property more broadly,[^2] economists have largely overlooked the dynamics of this period of "piracy" except in the service of broader arguments or economic history regarding intellectual property rights.
-@Khan2005, however, recognizes the idiosyncratic and endogenous component of copyright history, primarily to distinguish the unique and superior American system of IP from that of European systems.
-@Khan2004 examines its welfare effects on American authors, publishers, and consumers, finding it a net positive.
+Economists primarily focus on the public good aspects of ideas and expressions: once published, an idea can be utilized by many simultaneously, and that it becomes difficult for the initial producer to appropriate the value of their idea by excluding others (competitors and consumers) from using the idea without payment.
+One of the traditional economic arguments for copyright protection is that it makes such ideas, instantiated in books or other media, excludable — competitors and consumers risk legal liability for using protected works without the initial producer's consent.
+As such, publishers traditionally negotiate with authors to acquire copyright to their works, to enforce against rivals.
+@Khan2005 recognizes the idiosyncratic and endogenous nature of copyright history, including the period of U.S.
+literary piracy, primarily to distinguish the unique and superior American system of IP from that of European systems.
+@Khan2004 examines the welfare effects on American authors, publishers, and consumers of the systematic pirating of foreign works, finding it a net positive.
 @Spoo2013 [Ch.1] notably places the workings of the courtesy system in the literature of solutions to the tragedy of the commons, citing touchstone works such as @Ellickson1994 on neighbors solving disputes and @Rustiala2012 on norms in fashion, comedy, and cooking.
@@
 
-[^2]: See e.g. @Besen1991; @Posner2003 for a good survey. For a growing corpus of more critical works, see e.g. @Boldrin2008; @Kinsella2008; @Heller2008; @Rustiala2012; @Bell2014; @DouradoTabarrok2015
+[^2]: See e.g.
+@Besen1991; @Posner2003 for good surveys of the economic issues of intellectual property.
+For a growing corpus of more critical works, see e.g.
+@Boldrin2008; @Kinsella2008; @Heller2008; @Rustiala2012; @Bell2014; @DouradoTabarrok2015.
 
@@
 
-> "[Trade courtesy] was a brief realization of the ideals of philosophical anarchism--- self-regulation without law. ---Henry Holt (1908)2
+> "[Trade courtesy] was a brief realization of the ideals of philosophical anarchism--- self-regulation without law. ---Henry Holt (1908, 2)
 
@@
 Constitution[^3] [@Pattersonbook].
-Largely mimicking the 1710 Statute of Anne in Britain - the first modern copyright statute - it granted "sole right and liberty of printing, reprinting, publishing and vending" to authors of "any map, chart, book or books already printed within these United States" for "the term of fourteen years from the recording the title thereof in the clerk's office," with an optional 14-year renewal term (Copyright Act of 1790, \S 1).
+Largely mimicking the 1710 Statute of Anne in Britain — the first modern copyright statute — it granted "sole right and liberty of printing, reprinting, publishing and vending" to authors of "any map, chart, book or books already printed within these United States" for "the term of fourteen years from the recording the title thereof in the clerk's office," with an optional 14-year renewal term (Copyright Act of 1790, \S 1).
 Crucially for American publishers, section 5 of the Act explicitly denies copyright protection to foreigners:
 
-[^3]: "To promote the Progress of Science and useful Arts, by securing for limited Times to Authors and Inventors the exclusive Right to their respective Writings and Discoveries" (U.S.
+[^3]: "The Congress shall have Power...To promote the Progress of Science and useful Arts, by securing for limited Times to Authors and Inventors the exclusive Right to their respective Writings and Discoveries" (U.S.
 Constitution, Article I, \S8, Cl.
@@
 
-The domestic publishing industry in the early United States was largely run by Scottish and Irish immigrants such as Mathew Carey, who transplanted their tried-and-true experiences pirating London's books in Edinburgh and Dublin into New York and Philadelphia [@Johns2009, p.188ff; @Spoo2013, 34; @Groves2007b, 141].[^4] Early politicians and pundits saw reprinting as key to the emerging "American system" of political economy championed by Alexander Hamilton, Henry Clay, and Friedrich List, whereby tariffs and other measures would protect infant manufacturing industries.
-Thinkers of this school saw the reprinting (authorized or not) of European (predominantly British) works as key to instilling mass literacy, inspiration for innovation, instilling republican virtue [@Johns2009; @Spoo2013, 28; @Clark1960, 70].
+The domestic publishing industry in the early United States was largely run by Scottish and Irish immigrants such as Mathew Carey, who transplanted their tried-and-true experiences pirating London's books in Edinburgh and Dublin to New York and Philadelphia [@Johns2009, p.188ff; @Spoo2013, 34; @Groves2007b, 141].[^4] Early politicians and pundits saw reprinting as key to the emerging "American system" of political economy championed by Alexander Hamilton, Henry Clay, and Friedrich List, whereby tariffs and other measures would protect infant manufacturing industries.
+Such advocates saw the reprinting (whether authorized or not) of European (predominantly British) works as key to promoting mass literacy, inspiration for innovation, and instilling republican virtue [@Johns2009; @Spoo2013, 28; @Clark1960, 70].
 In the first few decades of the American colonies and early republic, the great distances between major cities kept parochial printing presses from cannibalizing each other's hold over their local market.
+New York publishers faced little competitive threat from Philadelphia or Boston-based publishing houses, and vice versa.
 
@@
 Harper & Brothers was the largest firm, with over 2 million volumes sold in 1859.
-
 The growth of the industry brought about fierce competition, especially after the Depression in the late 1830s and a ruinous price war in the 1840s.
@@
 This period of "cutthroat competition" caused publishers to recognize the need for self-regulation and organizational solutions within the industry into the 1840s.
-Some publishers, such as Carey, suggested forming an explicit organization modeled off of the London Stationers Company.
+Some publishers, such as Carey, suggested forming an explicit organization modeled off of the London Stationers Company, the state-sponsored guild that controlled and regulated publishing in Britain since the 16^th^ century (discussed below in Section 2.7).
 Carey set up The Philadelphia Company as an explicit cartel arrangement, only to see it evaporate 5 years later.[^6] Despite its failure, @Johns2009 [199] summarizes the common sentiment within the publishing industry:
@@
 
-[^6]: Once such a cartel allocated titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint.
-Without an inflexible cartel, a bargain could be reached between two publishers to establish boundaries, but this was not possible with a large organization and rogue outside printers.
+[^6]: Once such a cartel created a public Register to allocate exclusive titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint.
+Without an inflexible cartel, a bargain could be reached between two publishers to establish boundaries, as described below, but this possibility would be hindered by a large organization and rogue outside printers.
 
@@
   "c.1870s-1880s", "Second price war",
-  "1891", "Copyright Act grants foreign authors copyright protection in U.S."
+  "1891", "International Copyright Act grants foreign authors copyright protection in U.S."
 ) %>%
@@
 The foreign reprinting industry in the United States was spawned directly from Section 5 of the Copyright Act of 1790.
-An American publisher could publish works by *American* authors only by acquiring the copyright from them first.
+An American publisher could publish works by *American* authors only by negotiating for the copyright from them first.
 Since section 5 exempted *foreign* authors from copyright protection, it provided an indirect subsidy to reprint these works without concern for acquiring valid copyrights.
-As such, rather than paying and haggling with an *American* author for her copyright, an American publisher could print European authors at a significantly reduced cost.
-Both established and upstart publishers rapidly entered this market, searching for a quick opportunity to republish a "sure thing" to an American populace hungry for English literature.
+As such, rather than paying and haggling with an *American* author for her copyright, an American publisher could, in principle, print European authors at a significantly reduced cost.
+Both established and upstart publishers rapidly entered this market, searching for a quick opportunity to republish a "sure thing" to an American populace hungry for literature.
 
 The informal "courtesy of the trade" system of norms was both voluntary and ever-changing with market conditions.
-Not all publishers abided by trade courtesy, and some that followed norms closely for one decade would later violate them when it was to their benefit.
-Some firms that had cut their teeth in the profession by reprinting other publishers' reprints of foreign works later turned "legitimate" by following trade courtesy and enforcing it against upstart pirates.
+Not all publishers abided by trade courtesy, and some that upheld the norms closely for a decade would later violate them when it was to their benefit.
+Likewise, some firms that had cut their teeth in the profession by reprinting other publishers' reprints of foreign works later turned "legitimate" by following trade courtesy and enforcing it against new upstart pirates.
 
@@
 This was referred to in the trade by the "rule of association" [@Spoo2013,38].
-Aside from providing a bright line rule for priority of claims, a major benefit of the association rule was the a firm could advertise themselves as being the "authorized" publisher of a foreign author [@Spoo2013,42].
+Aside from providing a bright line rule for priority of claims, a major benefit of the association rule was that a firm could advertise themselves as being the "authorized" publisher of a foreign author [@Spoo2013,42].
 
-However, simply asserting that one *claimed* first priority in publishing an author was insufficient, and could incentivize "courtesy creep," [@Spoo2013,37]: Publishers might announce simultaneously or simply announce their intent to print any title that sounded promising, hoping to establish priority even if they would never print.
+However, simply asserting that one *claimed* first priority in publishing an author was insufficient, and could incentivize "courtesy creep," [@Spoo2013,37]: Publishers might announce simultaneously or simply announce their intent to print *any* title that sounded promising, hoping to establish priority even if they would never print it.
 Such a "first to announce" rule could quickly lead to inefficient rent-seeking: claimants merely assert claim in order to block others from using it, even if the claimant never intends to develop the resources in a productive way [@Posner2003,17].
@@
 
-Instead of a "first to announce rule," a "committed-searcher rule," where the first claimant to undertake costly steps (strongly correlated with actually developing the resource) improves upon these incentives [@Posner2003,17].
+Instead of a "first to announce rule," a "committed-publisher rule," where the first claimant to publicly undertake costly steps (strongly correlated with actually developing the resource) improves upon these incentives [@Posner2003,17].
 Publishers following trade courtesy required a firm seeking to stake a claim to undertake a more costly action, often reaching a public agreement with the foreign author, or more frequently, purchasing advance sheets of the edition abroad.
 Holt summarizes: "If a publisher had the advance sheets in his possession, such right or claim overrode a simple announcement," (quoted in @Harper1912, 110-11).
-@Spoo2013 [37] describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown, and the firm undertaking a risky experiment.
+@Spoo2013 [37] describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown.
+Holt describes that these cases were “a field for experiment, and if a publisher concluded one was worth experimenting with, though not worth paying for in advance of experiment, the rights from first announcement were intended to secure him the fruits of his experiment, if successful,” [@Publisherslibel, 360].
+In such cases of an unknown or new author, "property rights" over the exclusivity were relatively low in value, hence costly investments in securing exclusivity, such as purchasing advance sheets, were not required due to lack of competition.
+However, if the gamble of publishing a new author paid off and was a hit, the exclusive right to publish becomes more valuable, and that publisher had established priority through their actions.
 
@@
 
-To sustainably manage a commons, users must obtain benefits proportionate to the costs they bear.
+To sustainably manage a commons, users must obtain benefits proportionate to the costs they incur.
 Otherwise, if a substantial subset of the group free rides without contributing to the commons, or contributes without any reward, the tragedy of the commons may rear its ugly head, and the resource will be depleted.
-In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs.
+In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs, making it unprofitable for *anyone* to publish foreign uncopyrighted works.
 
@@
 
-Naturally, upstart publishers that were only in it for a quick buck had little incentive to worry about reputation.
+The logic of repeated dealings often was sufficient to constrain publishers from violating these norms.
+Consider a simple game where publisher $A$ and publisher $B$ (both of whom republish foreign authors in the United States), can *cooperate* with one another by upholding trade courtesy, or they can *defect* and print on each other.
+If both publishers uphold trade courtesy and cooperate, they each earn payoff $b$.
+If one cooperates and another defects, the cooperator earns payoff $d$ and the defector earns payoff $a$.
+If both defect, they each earn payoff $c$.
+In this simple prisoners’ dilemma, payoffs $a>b>c>d$.
+The game is summarized in Table 2.
+
+```{r, echo = F, warning = F, message = F}
+library(tidyverse)
+library(kableExtra)
+game <- tribble(
+  ~Player, ~Choice, ~Cooperate, ~Defect,
+  "Publisher A", "Cooperate", "b, b", "d, a",
+  "Publisher A", "Defect",  "a, d", "c, c"
+)
+game %>% 
+  mutate(Player = text_spec(Player, format = "latex")) %>%
+  mutate(Choice = text_spec(Choice, format = "latex")) %>% 
+  rename(` ` = Player, `  ` = Choice) %>% 
+  kbl(escape = FALSE, align = c("c", "c", "c", "c"), caption = "Payoffs a > b > c > d") %>% 
+  kable_styling(bootstrap_options = "basic", full_width = FALSE, position = "center") %>% 
+  collapse_rows(1) %>% 
+  add_header_above(c(" " = 2, "Publisher B" = 2), escape = FALSE)
+```
+
+In a one-shot game, each publisher has a dominant strategy to defect and pirate the work of other domestic publishers.
+Thus, both publishers earn $c$ in the Nash equilibrium.
+If the publishers could find a way to ensure that both cooperate, they would both be better off, each earning $b$.
+Trade courtesy and the attendant series of norms served primarily as this coordination device.
+Specifically, the logic of repeated dealings and the reputation that individual publishers cultivated as participants in this system constrained the incentive to defect.
+If this game is repeated, assuming publishers have a common future discount factor $\delta$, the folk theorem famously says that many equilibria are possible where both publishers cooperate (see e.g.
+@Axelrod1984; @DixitSkeath).
+Considering the limiting case of the "grim trigger strategy," where each publisher cooperates until the other defects, and then defects for all future turns, the threat of retaliation constrains individual publishers to cooperate so long as:
+
+$$\frac{b}{1-\delta}>a + \frac{\delta c}{1-\delta}$$
+i.e.
+so long as the discount factor is sufficiently high, $\delta > \frac{b-a}{c-a}$.
+Reputation serves as a depreciable capital asset, held hostage for a publisher's good behavior as a participant in this "gentlemanly" system of norms [@KleinAlchian; @Williamsonhostages].
+Once one becomes known as a "pirate," others will no longer respect claims to priority or association with any author.
+
+Naturally, upstart publishers that were only in it for a quick buck and then close up shop had little incentive to worry about reputation.
 So long as there were not too many publishers competing over price in this way, the industry could maintain this equilibrium.
@@
 Upholding proper "gentlemanly" conduct was not costless, it did require publishers to sacrifice profitable opportunities in the hope of not destroying their reputation (and hopes of future profit opportunities).
-As mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher *A* seeking to reprint works of foreign author *X* (and hoping to beat publisher *B* to the punch) could not merely assert this and hope publisher *B* would back down; publisher *A* needs to strike an exclusive arrangement with author *X* or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.
+As first mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher *A* seeking to reprint works of foreign author *X* (and hoping to beat publisher *B* to the punch) could not merely assert this and hope publisher *B* would back down; publisher *A* needs to strike an exclusive arrangement with author *X* or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.
 
-This required investment led to one of the great ironies of the system: despite the absence of a legal duty, U.S.
+This necessary investment led to one of the great ironies of the system: despite the absence of a legal duty, U.S.
 publishers frequently paid significant honoraria to British authors for the priority "right" to print their works.
@@
 Harper & Brothers, for example, paid Edward Bulwer-Lytton £50 per volume and Charles Dickens between £250 and £2,000 for his novels [@Spoo2013, 39].
-Harper Brothers, the largest 19<sup> th</sup> century publisher in the U.S.
+Harper Brothers, the largest 19<sup>th</sup> century publisher in the U.S.
 (and later the world) described the courtesy practice of purchasing advance sheets from foreign authors or their agents or publishers:
 
-> "In the absence of an international copyright, it is the custom for an English author, or his agent in London, to send early sheets to some American publisher, fixing a price therefor, and by a law of courtesy the American publisher who has issued the previous works of an author is entitled to the first consideration of that author's new book," [@Harper1912,358].
+> "In the absence of an international copyright, it is the custom for an English author, or his agent in London, to send early sheets to some American publisher, fixing a price therefore, and by a law of courtesy the American publisher who has issued the previous works of an author is entitled to the first consideration of that author's new book," [@Harper1912,358].
 
@@
 While many popular British authors publicly lamented the lack of copyright protection in America --- notably Charles Dickens, who made it a feature of his 1867 tour of the U.S.
---- they were often compensated even though publishers had no obligation to do so.
-Such honoraria and payment for the "exclusivity" that came with being first, or with the author's "authorized" version, was surely less than the author would have received in royalties or profit-sharing under copyright protection.
-British author Frederick Marryat estimated in 1839 that if Walter Scott (the British author who first became a best seller in the U.S.) lost out on \$500,000 in royalties from U.S.
+--- they were often well-compensated even though publishers had no obligation to do so.
+Honoraria and payments for the "exclusivity" that came with being first, or with the author's "authorized" version, was surely less than the author would have received in royalties or profit-sharing under full and legitimate copyright protection.
+British author Frederick Marryat estimated in 1839 that if Walter Scott (the first British author to became a best seller in the U.S.) lost out on \$500,000 in royalties from U.S.
 sales [@Madison1966,8-12].
@@
 To further underscore the copyright tradeoff between U.S.
-and Britain, Scott's novels often sold for 1/4 of the British price in the U.S., but gave him more American readers than English readers in total (ibid,12).
+and Britain, Scott's novels often sold for 1/4 of the British price in the U.S., but the demand was sufficiently elastic to yield him gave more American readers than English readers in total (ibid,12).
 
@@
 The *American Publishers' Circular and Literary Gazette* was so bold as to declare in 1872 that it was "an undeniable fact that there is no living English author of established reputation, whose books are extensively republished in this country, who is not freely and properly compensated by the American publisher," (quoted in @Sheehan1951 [68]).
-Publisher George Putnam more soberly explains that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," [@Putnam1915,365].
-By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of payments for advance sheets for popular authors [@Groves2007b,146].
+Publisher George Putnam more soberly lamented that "the author with no legal rights was thankful to get ten pounds when he could not get fifty and was very ready in receiving fifty to give a full quittance of any claim on the general proceeds," [@Putnam1915,365].
+By the end of the 19<sup>th</sup> century, a more formalized system of royalties on copies sold replaced the system of upfront payments for advance sheets for popular authors [@Groves2007b,146].
 
@@
 Some, like Henry Carey, not-so-secretly dreamed of organizing an American guild like the Stationers Company back in England.[^7] However, such a scheme always failed because it lacked the flexibility of the courtesy system in two important ways [@Johns2009,197-202]: First, such a corporation would generate a list of titles (akin to the Register of the Stationers Company) available only to its members --- boosting a signal to rogue reprinters that these titles were "sure things" worth printing on.
-Second, a formal corporation prevented the very bargains and exchanges between publishers to settle psuedo-property rights disputes to prevent American publishers printing on each other.
+Second, a formal corporation prevented the very bargains and exchanges between publishers to settle pseudo-property rights disputes to prevent American publishers printing on each other.
 
@@
 While trade courtesy norms could not sustain themselves without vigorous monitoring, each individual publisher had a strong incentive to monitor others for violating norms.
-If a publisher's works were printed on by other publishers, that publisher stood to lose profits from competition.
-Thus, each individual publisher had an interest in ensuring that other publishers respected their claims, and via reciprocity in this manner, the system of trade courtesy was largely self-enforcing.
+If a publisher's works were printed on by other publishers, that publisher stood to directly lose profits from competition.
+Thus, each individual publisher had an interest in policing their claims, and via reciprocity in this manner, the system of trade courtesy was largely self-enforcing.
 However, it is still instructive to see how publishers commonly took several types of actions in order to notify others of their claims and dispute the claims of others in a public manner.
@@
 This network proved critical for communicating about claims, gossiping about others' claims, resolving honest disputes, and calling out transgressors.
-Publishers regularly wrote to their customers, advertised in newspapers, and created a number of trade journals to circulate announcements, reviews, and notices to one another [@Groves2007a].
+Publishers also regularly wrote to their customers, advertised in newspapers, and created a number of trade journals to circulate announcements, reviews, and notices to one another [@Groves2007a].
 Periodicals such as *Publishers' Weekly*, which by the 1880s emerged dominant within the trade, also advocated consensus positions within the trade for regulatory reform.
 
-Beyond the frequent communication between individual publishers, it was essential to maintain the nexus of assertions of ownership and association with foreign authors publicly, by posting various messages in widely-read trade journals.
+Beyond the frequent communication between individual publishers, it was essential to maintain the nexus of assertions of ownership and association with foreign authors publicly by posting various messages in widely-read trade journals.
 In the 1850s, there was no agreed upon single medium, as various publishers utilized different trade journals and weeklies.
-Due to the lack of a single medium, publishers frequently clipped and kept records of their own advertisements in media in order to provide justification to their claims should disputes later arise @Groves2007b, 142].
-As the practice matured, publishers eventually reached consensus that claims "in press" should be advertised and declared in the *New York Commercial Advertiser* [@Groves2007b, 146; @Spoo2013, 37; @Sheehan1951, 69].
+Due to the lack of a single medium, publishers frequently clipped and kept records of their own advertisements in media in order to provide justification for their claims should disputes later arise [@Groves2007b, 142].
+As the practice matured, publishers eventually reached consensus that claims "in press" should be advertised and declared specifically in the *New York Commercial Advertiser* [@Groves2007b, 146; @Spoo2013, 37; @Sheehan1951, 69].
 Other major trade journals such as *The Publishers' Weekly* and *The Weekly Trade Circular* then reprinted the advertisements from the *New York Commercial Advertiser* to ensure sufficient dissemination that publishers in other cities were properly warned not to print on the claim.
@@
 
-> "We notice in last evening's Commercial Advertiser your 'in press' announcement of the 'Life of Marie Antoinette' by C.
+> "We notice in last evening's *Commercial Advertiser* your 'in press' announcement of the 'Life of Marie Antoinette' by C.
 D.
@@
 An earnest dispute over who had priority might first be met with appeals to publishers' reputation or public shaming.
-@Spoo2013 [43] cites an example in 1870 where the Harpers attempted to print on Holt's edition of Hippolyte Taine's *On Intelligence*.
+@Spoo2013[43] cites an example in 1870 where the Harpers attempted to print on Holt's edition of Hippolyte Taine's *On Intelligence*.
 Holt wrote to the Harpers merely asking "Doesn't the fact that we have published several of [Taine's] books entitle us to [exclusively publishing *On Intelligence*] if we want it?" Such a simple response was sufficient for the Harpers to back off, abiding by the rule of association.
@@
 Firms that were associated with foreign authors might protect their claim against other publishers printing on them by including copies of letters penned by the foreign author at the beginning of the book.
-Ticknor & Fields' editions of Thomas De Quincey's works contained a letter featuring the latter approving of the former "exclusively" publishing his works in America.
+Ticknor & Fields' editions of Thomas De Quincey's works contained a letter featuring the author approving of the publisher "exclusively" publishing his works in America.
 Similarly, United States Book Co.
@@
 Ironically, one of the most effective forms of shaming was to denounce transgressing publishers as "pirates" before their more respectable peers.
-At a libel lawsuit reported by *Publisher's Weekly*, Holt and other major publishers testified that use of the term "pirate" had a special meaning in American publishing - as a public denunciation within the profession for those that refused to abide by trade courtesy [@Publisherslibel].
-If a publisher were recalcitrant to such public denunciations, publishers could rally together for other multilateral punishments, such as refusal to deal with a rogue publisher.\
-\
-The ultimate sanction against a "pirate", which if not kept in check by agreement that its *occasional* use was legitimate, was retaliation by predatory pricing.
+At a libel lawsuit reported by *Publisher's Weekly*, Holt and other major publishers testified that use of the term "pirate" had a special meaning in American publishing — as a public denunciation within the profession for those that refused to abide by trade courtesy [@Publisherslibel].
+If a publisher were recalcitrant to such public denunciations, publishers could rally together for other multilateral punishments, such as refusal to deal with a rogue publisher.
+
+The ultimate sanction against a recalcitrant publisher, utilized only as a last resort, was retaliation by predatory pricing.
 While the *raison d'etre* of trade courtesy was to keep prices sufficiently high to avoid a price war, publishers recognized it as a legitimate response or a preemptive strike against pirates.
@@
 
-Printing on other publishers' claims ran frequent during the two price wars of the 19<sup>th</sup> century.
+Printing on other publishers' claims ran frequent during the two price wars of the 19<sup>th</sup> century in the 1840s and 1870s.
 In the 1870s, for example, Harper Brothers reprinted their previously published foreign novels at just 10 cents a piece.
@@
 
-Retaliation, if not kept in check, could easily spin out of control.
+Such retaliation, if not kept in check, could easily spin out of control.
+The "property right" over an exclusive work becomes worthless if it is no longer exclusive and the price of the work is competed down to cost, or even below it.
 When Harper & Brothers failed to resolve a dispute over Tennyson's works with Fields, Osgood & Company --- who had been associated with the author for 30 years --- in 1870, Harper released their own edition [@Sheehan1951,145].
@@
 Much like the owner of a true copyright today, a publisher with a good claim to issuing a foreign work in the U.S.
-could bargain with other publishers over a wide range of in-kind rights - issuing serials, hardbound books, paperback books, etc.
-- to reach an amicable arrangement.
+could bargain with other publishers over a wide range of in-kind rights — issuing serials, hardbound books, paperback books, etc.
+— to reach an amicable arrangement.
 
@@
 Trade courtesy in publishing was primarily a response to explicit legal institutions, particularly the gap in copyright protection between domestic and foreign authors.
-Historical trade practices in printing, since the dawn of the printing press, had been shaped by changes in legal and political institutions.
+Historical trade practices in printing, since the dawn of the printing press, has been shaped by changes in legal and political institutions.
 Since the first printers in England in 1476, printers had tried to organize themselves internally to manage the commons of "printing on" each other's works, regardless of whether the legal environment recognized explicit property rights in works.
@@
 While often politically controversial in the tumultuous English Civil Wars of the 17^th^ Century [@Safnerpatronage], the main feature of the Stationers' practices was the establishment of a Register that maintained who had the exclusive, perpetual, and transferable right to print a particular work (known as "a copy right") as well as a series of sanctions and methods of adjudicating disputes based on copies authorized by the Register.
-Alongside this grew up "trade customs" of printing as an honorable craft.
+Alongside this formal organization grew up more informal norms of "trade custom" that maintained the dignity of printing as an honorable craft.
 The passage of the Statute of Anne in 1710 created the first statutory copyright in th U.K.
 (a temporary legal monopoly), but left unaddressed the status of the Stationers' copyright until the 1774 case of *Donaldson v.
-Beckett*, when the House of Lords decided that literary expression is protected only by statutory copyright.[^9] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy of famous works out of London.
+Beckett*, when the House of Lords decided that literary expression is protected *only* by the statutory copyright which Parliament had created almost *ex nihilo*.[^9] Prior to the dominance of statutory law governing the trade in the United Kingdom, printers and booksellers in Scotland and Ireland formed their own practices and customs regarding the systematic piracy and unauthorized distribution of famous works out of London to the peripheries of the Kingdom.
 As noted above, many of these tradesmen emigrated to the early United States to carry on their esteemed tradition of pirating English works.
 
-[^9]: The Stationers conjured a clever rhetorical idea that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the protections of the Statute of Anne, a legal claim the Lords definitely rejected, @Johns2009.
+[^9]: The Stationers conjured a clever rhetorical idea that their internal trade practices had constituted a "common law" copyright that existed prior to and separate from the statutory protections of the Statute of Anne, a legal claim the Lords definitively rejected, [@Johns2009].
 
-The courtesy of the trade was only one aspect of a whole political economy of the young republic, a significant part of which revolved around protecting American publishers and readers.
-Many argued, especially in the early 19^th^ Century for a new "American system" of mercantilism favorable to domestic manufacturing in order to promote national development.
+The courtesy of the trade was only one aspect of a whole political economy of the young American republic, a significant part of which revolved around protecting domestic publishers and readers.
+Many American political figures argued, especially in the early 19^th^ Century for a new "American system" of mercantilism favorable to domestic manufacturing in order to promote national development.
 The "appropriation" of European literature and technology was encouraged as a stimulant to American development and even instilling republican virtue [@Johns2009, 203].
 A gradually-increasing tariff (ranging from 10-25%) slowed the importation of foreign books starting in 1816 and implicitly subsidized the American publishing (and reprint) industry [@Groves2007b, 141; @Dozer1949, 73].
-@Dozer1949 [95] argues that the issue of respecting foreign copyright in America "has been...a tariff question involving the protection of American manufacturing interests." A federal court also recognizes the 1909 Copyright Act and a later Tariff Act (of 1930) as sharing a common purpose, "to encourage the industries of the United States," and "to protect American labor," (*Oxford University Press v.
+@Dozer1949 [95] argues that the issue of respecting foreign copyright in America explicitly "has been...a tariff question involving the protection of American manufacturing interests." A federal court also recognizes the 1909 Copyright Act and a later Tariff Act (of 1930) as sharing a common purpose, "to encourage the industries of the United States," and "to protect American labor," (*Oxford University Press v.
 United States*, 1945 (No.
@@
 
-Of course, while the courtesy of the trade was able to often replicate the features of property rights and even copyright law itself, such voluntary norms were not viewed legitimate or enforceable property rights in court.
+Of course, while the courtesy of the trade was able to often replicate the features of property rights and even copyright law itself, such voluntary norms were not viewed as legitimate or enforceable property rights in court.
 The federal court in the 1865 case of *Sheldon v.
@@
 > [trade courtesy] confessedly rests upon no common law of the country, recognized and administered by judicial tribunals.
-If it has any foundation at all, it stands on the mere will, or...the "courtesy" of the trade...It can, therefore, hardly be called property at all---certainly not in any sense known to the law, quoted in [@Spoo2013, 34].
+If it has any foundation at all, it stands on the mere will, or...the "courtesy" of the trade...It can, therefore, hardly be called property at all---certainly not in any sense known to the law, quoted in @Spoo2013[34].
 
 The system rapidly broke down after 1891 with the passage of the International Copyright Act.[^11] The Act recognized foreign copyrights in the United States provided that the works themselves were manufactured in the United States.
-This notorious "manufacturing clause," as well as other qualifications and requirements kept the United States ineligible to join the 1886 Berne Convention until 1989, following major revisions to U.S.
+This notorious "manufacturing clause," as well as other qualifications and requirements kept the United States ineligible to join the 1886 Berne Convention (an international treaty standardizing the recognition of copyrights, including foreign copyrights) until 1989, following major revisions to U.S.
 law.
@@
 Ezra Pound's lament of the "thieving copyright law" as late as the 1920s was representative of European attitudes towards the sole great power not party to the Berne Convention [@Spoo1998, 645].
-While the passage of the copyright law is certainly endogenous to the market conditions facing the publishing industry [@Safnerpiratethyneighbor], it is beyond the scope of this paper to discuss both the causes and effects of the legal change.\
-\
+While the passage of the copyright law is certainly endogenous to the market conditions facing the publishing industry [@Safnerpiratethyneighbor], it is beyond the scope of this paper to discuss both the causes and effects of this major legal change.
+
 Regardless of the legal imperfections of international copyright, the courtesy of the trade system that had organized "honorable" American piracy of foreign works served little purpose following the 1891 law.
-While a form of trade courtesy persisted and evolved among American publishers into the mid-20^th^ century, it had little in common with that of its 19^th^ century progenitor.
+While a variant of trade courtesy persisted and evolved among American publishers into the mid-20^th^ century, it had little in common with that of its 19^th^ century progenitor.
 Royalties replaced payments for advance sheets, and @Sheehan1951 [79] finds that foreign authors seem to have obtained about the same terms with American publishers as did American authors.
 For example, Charles Scribner's Sons' royalties on 10 contracts between 1901-1906 seem to have varied between 10-20% for foreign authors and domestic authors alike (ibid).
-The rule of association, however, was extended to the burgeoning class of American authors, ensuring that publishers would not attempt to poach an author from her associated publisher, unless the author made the first move [@Groves2007b, 141].
+The rule of association, however, was now extended to the burgeoning class of American authors, ensuring that publishers would not attempt to poach an author from her associated publisher, unless the author made the first move [@Groves2007b, 141].
 
@@
 In 1886 alone, 26 "cheap libraries" issued 1,500 different titles, mostly novels priced between 10-50 cents, thinning the industry's profit margins [@Spoo2013,53].
-Appealing to populist notions of cheap literature for the masses, these entrants directly attacked courtesy as a price-fixing cartel.
+Appealing to populist notions of cheap literature for the masses, these entrants denounced trade courtesy as thinly-veiled collusion among incumbent publishers.
 Lovell opined in an 1879 edition of *Publishers' Weekly* that:
@@
 
-@Sheehan1951 [24] suggests that the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^12] The largest firm, Harper & Brothers, by far, constituted less than 2% of the industry's sales in 1914.\
-\
+@Sheehan1951 [24] suggests that despite the rise in anti-cartel and anti-trust rhetoric, both within the publishing trade and in the U.S.
+more broadly, the limited evidence available suggests no monopolistic tendency in the publishing industry over this time period.[^12] The largest firm, Harper & Brothers, by far, comprised less than 2% of the industry's sales in 1914.
+Indeed, the entry of the new cheap publishers and the concomitant price wars kept the industry fiercely competitive.
+
 Furthermore, even those firms that upheld trade courtesy knew it was imperfect, and was little more than a paper shield against determined pirates.
@@
 Even a few days or hours made all the difference between a profit or a pirate-induced loss on a venture.
-Harper Brothers, for example, could frequently outrun its largest competitor, Carey & Lea shipping from New York into the Ohio valley, and at half the cost, compared with Philadelphia @Madison1966, 12,21].
+Harper Brothers, for example, could frequently outrun its largest competitor, Carey & Lea shipping from New York into the Ohio valley, and at half the cost, compared with Philadelphia [@Madison1966, 12,21].
 
@@
 First, it suggests that even in the absence of formal property rights protections on expressions such as written works, it is in the interest of the producers to find ways to protect their property.
-By treating published works as a commons, always with the tragedy of the commons lurking in the background, the creative industries (in this case, book publishers) share a common interest in not depleting it.
+By treating published works as a commons, always with the looming threat of tragedy, the creative industries (in this case, book publishers) share a common interest in not depleting it.
 Many of the solutions to this problem replicate formal property rights, contracts, and copyright protection by creating exclusivity, the duty to respect others' exclusivity, and a set of remedies for violations.
@@
 On the one hand, the system allowed a number of American publishers to maintain a large degree of exclusivity over their products to protect profits, but on the other hand, as a voluntary system, the ever-present threat, and occasional intrusion of competition by pirates kept the market contestable, and prices low for consumers.
-While foreign authors were explicitly denied equal rights on par with their own homeland, or with American authors, the need publishers to establish claims of association led to them compensating foreign authors even when there was no legal duty to do so.
+While foreign authors were explicitly denied equal rights on par with their own homeland, or with American authors, the need of publishers to establish claims of association led to them compensating foreign authors even when there was no legal duty to do so.
 
-The notion of intellectual property itself remains contentious, with some claiming it is more the result of rent-seeking than a legitimate "right" akin to tangible property [@Boldrin2008; @Kinsella2008; @Bell2014].
+The scope and nature of of intellectual property itself remains contentious, with some claiming it is more the result of rent-seeking than a legitimate "right" akin to tangible property [see e.g.
+@Boldrin2008; @Kinsella2008; @Bell2014].
 Regardless of one's normative view of copyright, it remains in the interest of authors and especially publishers to establish property rights over printed works, whether formal or informal.
